---
title: "FY2026 Revenue Report"
format:
  html:
    theme: zephyr
    toc: true
    toc-location: left
    code-fold: true
    code-tools: true
    code-overflow: wrap
    code-copy: true
    df-print: paged
---


```{r}
#| label: setup
#| warning: false
#| message: false


knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(tidyverse)
library(formatR)
library(lubridate)
library(scales)
library(kableExtra)
library(ggplot2)
library(readxl)
library(janitor)
library(cmapplot)

alea_theme <- function() {
  font <-"Whitney"
  
  ggplot2::theme(
    legend.position = "right",
    legend.title = element_blank(),
    
    panel.background = ggplot2::element_blank(),
    panel.grid.minor.x = ggplot2::element_blank(),
    panel.grid.major.y = element_line(color = "grey"),
    panel.grid.minor.y = element_line(color = "grey", 
                                      linetype = "dashed"),
    # panel.grid.major.x = ggplot2::element_blank(),
    axis.ticks = element_line(color = "gray"),
    axis.ticks.x = element_blank()
  )
  
}

theme_set(alea_theme())

# Custom billion format
label_billions <- function(digits = 1) {
  function(x) {
    number_format(accuracy = 10^-digits, suffix = "B")(
      x / 1e9
    )
  }
}

scale_y_billions <- function(..., digits = 1) {
  scale_y_continuous(labels = label_billions(digits), ...)
}

move_to_last <- function(df, n) df[c(setdiff(seq_len(nrow(df)), n), n), ]


current_year <- 2026 

# fiscal year, not calendar year
past_year=current_year-1

rev_temp <- read_csv(paste0("../../Fiscal Futures IGPA/Fiscal-Future-Topics/data/FY", current_year, " Files/rev_temp.csv")) |> 
  mutate(fund = str_pad(fund, width = "4", side = "left", pad = "0"))

exp_temp <- read_csv(paste0("../../Fiscal Futures IGPA/Fiscal-Future-Topics/data/FY", current_year, " Files/exp_temp.csv")) |> 
  mutate(fund = str_pad(fund, width = "4", side = "left", pad = "0"))

```

```{r}
#| include: false


update_recessions <- function(url = NULL, quietly = FALSE){

  # Use default URL if user does not override
  if (is_null(url) | missing(url)) {
    url <- "https://data.nber.org/data/cycles/business_cycle_dates.json"
  }
  
   # locally bind variable names
  start_char <- end_char <- start_date <- end_date <- ongoing <- index <- peak <- trough <- NULL

  return(
    # attempt to download and format recessions table
    tryCatch({
      recessions <- jsonlite::fromJSON(url) |>
        # drop first row trough
        dplyr::slice(-1) |>
        # convert peaks and troughs...
        dplyr::mutate(
          # ...to R dates
          start_date = as.Date(peak),
          end_date = as.Date(trough),
          # ... and clean char strings
          start_char = format(start_date, "%b %Y"),
          end_char = format(end_date, "%b %Y")) |>
        # confirm ascending and create row number
        dplyr::arrange(start_date) |>
        mutate(index = row_number()) |>
        mutate(
          # Flag unfinished recessions
          ongoing = case_when(
            is.na(end_date) & index == max(.$index) ~ T,
            TRUE ~ F),
          # set ongoing recession to arbitrary future date
          end_date = case_when(
            ongoing ~ as.Date("2200-01-01"),
            TRUE ~ end_date),
          # mark ongoing recession in char field
          end_char = case_when(
            ongoing ~ "Ongoing",
            TRUE ~ end_char)
          ) |>
        # clean up
        select(start_char, end_char, start_date, end_date, ongoing)

      if (!quietly) {message("Successfully fetched from NBER")}

      # Return recessions
      recessions
    },
    error = function(cond){
      if (!quietly) message("WARNING: Fetch or processing failed. `NULL` returned.")
      return(NULL)
    }
    )
  )
}

recessions <- update_recessions()

```


# Step by Step Data Exploration and Table/Figure Building

## Modify Expenditure File


### Tax refunds


Aggregate expenditures: Save tax refunds as negative revenue. Code refunds to match the rev_type codes 


- 02 = income taxes, fund 0278  
- 03 = corporate income taxes,  fund 0946, 0380   
- 06 = sales tax  
- 09 = motor fuel tax, fund 0436, 0012  
- 24 = insurance taxes and fees  
- 35 = all other tax refunds.   

    - Estate tax refund, fund 0121  
    - Aviation fuel refund fund, fund 0946     
    - cannabis Excise Tax Refund, fund 0912     
    - Rental Purchase agreement Refund 0671   


What we want:   

- To exclude refunds as expenditures from our expenditure totals.  
- Exclude funds that hold refund funds from both revenue and expenditure sides  
- Revenue neutral unless budget tricks are occurring  
- still want to examine refunds over time to compare to past years calculations  



```{r}
#| label: fig-tax-refunds
#| fig-cap: "Tax Refunds" 


tax_refund_long <- exp_temp |>           # fund != "0401" # removes State Trust Funds
  filter(fund != "0401" &
           (object == "9900" | object=="9910"|object=="9921"|object=="9923"|object=="9925")) |>
  # keeps these objects which represent revenue, insurance, treasurer,and financial and professional reg tax refunds
  mutate(refund = case_when(
    object == "9900" & fund == "0278" ~ "FY23_Rebates",
    fund=="0278" & sequence == "00" ~ "02", # for income tax refund
    fund=="0278" & sequence == "01" ~ "03", # tax administration and enforcement and tax operations become corporate income tax refund
        fund=="0380" ~ "03", # corporate franv tax refund

     fund == "0278" & sequence == "02" ~ "02",
    object=="9921" ~ "21",                # inheritance tax and estate tax refund appropriation
    object=="9923" ~ "09",                # motor fuel tax refunds
    obj_seq_type == "99250055" ~ "06",    # sales tax refund
    fund=="0378" & object=="9925" ~ "24", # insurance privilege tax refund
   (fund=="0001" & object=="9925") | (object=="9925" & fund == "0384" & fy == 2023) ~ "35", # all other taxes
   # fund=="0001" & object=="9925" ~ "35", # all other taxes
       fund %in% c("0946", "0912", "0671")  ~ "35", # cannabis, aviation, rental purchase tax refund
    T ~ "CHECK"))                       # if none of the items above apply to the observations, then code them as CHECK 

    
exp_temp <- left_join(exp_temp, tax_refund_long) |>
  mutate(refund = ifelse(is.na(refund),"not refund", as.character(refund)))

tax_refund <- tax_refund_long |> 
  group_by(refund, fy)|>
  summarize(refund_amount = sum(expenditure, na.rm = TRUE)) |>
  pivot_wider(names_from = refund, values_from = refund_amount, names_prefix = "ref_") |>
  mutate_all(replace_na, 0) |>
  arrange(fy)

# tax_refund |>
#   pivot_longer(c(ref_06:ref_35, ref_FY23_Rebates), names_to = "Refund Type", values_to = "Amount") |>
#   ungroup() |>
#   ggplot()+
#   geom_line(aes(x=fy, y=Amount, group = `Refund Type`, color = `Refund Type`))+
#     scale_y_billions() +
#   labs(title = "Refund Types") +
#   labs(title = "Tax refunds",
#        caption = "Rev_type codes: 02=income taxes, 03=corporate income taxes, 06=sales tax, 09=motor fuel tax,
#        24=insurance taxes and fees, 35 = all other tax refunds.",
#        y="Dollars", x = element_blank()
#        ) 

```


```{r}
#| label: fig-tax-refunds-noabatement
#| fig-cap:  "Tax Refunds without FY23 Abatements"

tax_refund_long <- exp_temp |>           # fund != "0401" # removes State Trust Funds
  filter(fund != "0401" &
           (object == "9900" | object=="9910"|object=="9921"|object=="9923"|object=="9925")) |>
  # keeps these objects which represent revenue, insurance, treasurer,and financial and professional reg tax refunds
  mutate(refund = case_when(
    object == "9900" & fund == "0278" ~ "FY23_Rebates",
    fund=="0278" & sequence == "00" ~ "02", # for income tax refund
    fund=="0278" & sequence == "01" ~ "03", # tax administration and enforcement and tax operations become corporate income tax refund
        fund=="0380" ~ "03", # corporate franv tax refund

     fund == "0278" & sequence == "02" ~ "02",
    object=="9921" ~ "21",                # inheritance tax and estate tax refund appropriation
    object=="9923" ~ "09",                # motor fuel tax refunds
    obj_seq_type == "99250055" ~ "06",    # sales tax refund
    fund=="0378" & object=="9925" ~ "24", # insurance privilege tax refund
   (fund=="0001" & object=="9925") | (object=="9925" & fund == "0384" & fy == 2023) ~ "35", # all other taxes
   # fund=="0001" & object=="9925" ~ "35", # all other taxes
       fund %in% c("0946", "0912", "0671")  ~ "35", # cannabis, aviation, rental purchase tax refund
    T ~ "CHECK"))                       # if none of the items above apply to the observations, then code them as CHECK 

    
exp_temp <- left_join(exp_temp, tax_refund_long) |>
  mutate(refund = ifelse(is.na(refund),"not refund", as.character(refund)))

tax_refund <- tax_refund_long |> 
  group_by(refund, fy)|>
  summarize(refund_amount = sum(expenditure, na.rm = TRUE)) |>
  pivot_wider(names_from = refund, values_from = refund_amount, names_prefix = "ref_") |>
  mutate_all(replace_na, 0) |>
  arrange(fy)
 
tax_refund |>
  pivot_longer(c(ref_06:ref_35), names_to = "Refund Type", values_to = "Amount") |>
  ggplot()+
  theme_classic()+
  geom_line(aes(x=fy,y=Amount, group = `Refund Type`, color = `Refund Type`))+
  labs(title = "Refund Types") +
  labs(title = "Tax refunds without FY23 Abatements",
       caption = "Rev_type codes: 02=income taxes, 03=corporate income taxes, 06=sales tax, 09=motor fuel tax,
       24=insurance taxes and fees, 35 = all other tax refunds.",
       ) +
  scale_y_billions()

```



```{r}
#| label: fig-includedrefunds
#| fig-cap: "Based on current fund coding in the fund_ab_in.xlsx file, most refund dollars are already being excluded by having `in_ff` = 0. Funds that have `in_ff` = 0 are excluded from revenue and expenditure sums when calculating the Fiscal Gap."

tax_refund_long |> 
  summarize(expenditure = sum(expenditure, na.rm=TRUE), .by = c(fy, in_ff) ) |>
  ggplot() + 
  geom_line(aes(x=fy, y = expenditure, group = factor(in_ff), color = factor(in_ff)))+
  labs(title = "Excluded Refund Expenditures", y = "Dollars", x = element_blank())

```


```{r}
#| label: fig-tax-refunds-revenueside
#| fig-cap: "Tax Refunds from Revenues. Includes only funds designated as Refund funds in fund descriptions."

tax_refund_long_rev <- rev_temp |>          
  mutate(refund = case_when(
    fund == "0121" ~ "35", # Estate tax refund
    fund=="0278" ~ "02", # for income tax refunds (individual and corporate)
    fund=="0380" ~ "03", # corporate franchise tax refund
    fund=="0378" ~ "24", # insurance privilege tax refund
    fund %in% c("0946", "0912", "0671")  ~ "35", # cannabis, aviation, rental purchase tax refund
    T ~ "CHECK")) |>                      # if none of the items above apply to the observations, then code them as CHECK 
  filter(refund != "CHECK")

tax_refund_rev <- tax_refund_long_rev |> 
  group_by(refund, fy)|>
  summarize(allocated_for_refunds = sum(receipts, na.rm = TRUE)/1000000) |>
  pivot_wider(names_from = refund, values_from = allocated_for_refunds, names_prefix = "ref_") |>
  mutate_all(replace_na, 0) |>
  arrange(fy)
 
tax_refund_rev|>
  pivot_longer(c(ref_02:ref_35), names_to = "Refund Type", values_to = "Amount") |>
  ggplot()+
  geom_line(aes(x=fy,y=Amount, group = `Refund Type`, color = `Refund Type`))+
  labs(title = "Refund Types") +
  labs(title = "Revenue Allocated to Tax Refund Funds ",
       caption = "Rev_type codes: 02=income taxes, 03=corporate income taxes, 06=sales tax, 09=motor fuel tax,
       24=insurance taxes and fees, 35 = all other tax refunds.",
       y = "Millions of $", x = element_blank()) + 
    scale_x_continuous(expand = c(0,0), limits = c(1998, current_year+.5), breaks = c(1998, 2005, 2010, 2015, 2020, current_year))


```


Abatements were a specific policy choice to do and different than normal refunds. Therefore, abatements were included in the fiscal gap calculation. This is not even a concern during most years. Only was an issue in 2023 when $1 billion abatement expenditure occurred. 



```{r}

# manually adds the abatements as expenditure item and keeps on expenditure side.
# otherwise ignored since it is in fund 0278, which is coded as in_ff=0
# all other income tax refunds are excluded from fiscal gap calculations

exp_temp <- exp_temp |>
  mutate(in_ff = ifelse(object == 9900 & fund == "0278", 1, in_ff))


```



### Pension Expenditures


**An Explanation of a Methodological Change in How We Categorize Some Pension Spending**

In previous reports, a expenditure category was created to represent all Pension spending. It was created by removing pension expenditures out of the agencies that that had the expenses in order to highlight the amount that was spent on pensions.

As of the FY2024 report, pension expenditures will remain in the agency that has the pension expenses to better represent the total cost of providing a service to the public.

However, we thought that readers also might be interested in total pension spending during current and past years. We have added a separate table showing this information but emphasize that pension spending is already incorporated in other spending categories and should not be added to the total shown in Table 1 of the report.



New POB bond in 2019: Accelerated Bond Fund paid benefits in advance as lump sum   


State pension contributions for TRS and SURS are largely captured with object=4431. **(State payments into pension fund).** State payments to the following pension systems:

-   Teachers Retirement System (TRS) Agency 593 -->    K-12 Education (Group = 959)
-   State Employee Retirement System (SERS) Agency 589 -->    SERS Agency 589 - _Note: Object 4431 does not have SERS expenditures in it. Those are only in object 116X objects_   
-   State University Retirement System (SURS) Agency 693 -->  University Education (Group = 960)   
-   Judges Retirement System (JRS) Agency 275  --> Judicial (Group = 920)   
-   General Assembly Retirement System (GARS)  --> Legislative (Group 910)    


There are also "Other Post-Employment Benefits" (OPEBs). Expenditure object 4430 is for retirement benefits.

While it is good to know the overall cost of pensions for the state, if you want to know the true cost of providing services, pension and other benefit costs should be included in the department that is paying employees to provide those services. 



But if you are curious how much went into pensions and for which pension agency, they are totaled below: 

```{r}
#| layout-ncol:  2

pension_agencies <- c("589", "593", "594", "693", "275", "131" )

pension_objects <- c(4431, 1160:1165)

exp_temp |> 
  filter(
     fy == 2024 &
       object %in% pension_objects) |>
  mutate(group = ifelse(!agency %in% pension_agencies, "901", agency),
         group_label = case_when(
           group == "593" | group == "594" ~ "TRS",
           group == "589" ~ "SERS",
           group == "693" ~ "SURS",
           group == "275" ~ "JRS",
           group == "131" ~ "GARS",
          group == "901" ~ "Other Pension Costs",

           
           TRUE ~ "CHECK ME")) |>
 #   group = ifelse(object %in% 1160:1165, "901", as.character(agency))) |> 
  group_by(group, group_label) |>
  summarize(Expenditures = sum(expenditure, na.rm=TRUE)/1000000000)
 

exp_temp |> 
  filter(
     fy == 2025 &
       object %in% pension_objects) |>
  mutate(group = ifelse(!agency %in% pension_agencies, "901", agency),
         group_label = case_when(
           group == "901" ~ "Other Pensions",
           group == "593"  ~ "K-12 Education",
           group == "594" ~ "Chi. TPF ",
           group == "589" ~ "SERS",
           group == "693" ~ "Higher Education",
           group == "275" ~ "Judicial",
           group == "131" ~ "Legislative",
           
           TRUE ~ "CHECK ME")) |>
 #   group = ifelse(object %in% 1160:1165, "901", as.character(agency))) |> 
  group_by(group, group_label) |>
  summarize(Expenditures = sum(expenditure, na.rm=TRUE)/1000000000)

exp_temp |> 
  filter(
     fy == 2024 &
       object %in% pension_objects) |>
  mutate(group = ifelse(!agency %in% pension_agencies, "901", agency),
         group_label = case_when(
           group == "901" ~ "Other Department Pensions",
           group == "593" ~ "K-12 Education",
           group == "594" ~ "Chi. TPF ",
           group == "589" ~ "Other Department Pensions",
           group == "693" ~ "Higher Education",
           group == "275" ~ "Judicial",
           group == "131" ~ "Legislative",
           
           TRUE ~ "CHECK ME")) |>
 #   group = ifelse(object %in% 1160:1165, "901", as.character(agency))) |> 
  group_by(group, group_label) |>
  summarize(Expenditures = sum(expenditure, na.rm=TRUE)/1000000000)

# in billions 
exp_temp |> 
  filter( fy == 2025 &
            agency!="494" &
            object %in% pension_objects ) |>
  summarize(`Pension Expenditures` = sum(expenditure/1000000000, na.rm=TRUE))

```

_**TO DO: Relabel image and 2004 pension spike**_

```{r}
pension_totals <-  exp_temp %>% 
  arrange(fund) %>%
  mutate(pension = case_when( 
  
            (object=="1298" &  # Purchase of Investments, Normally excluded
       (fy==2010 | fy==2011) & 
       (fund=="0477" | fund=="0479" | fund=="0481")) ~ 3,  
       (object=="4431") ~ 1, # 4431 = easy to find pension payments INTO fund

    
     (object>"1159" & object<"1166") & fund != "0183" & fund != "0193"   ~ 2, 
    # objects 1159 to 1166 are all considered Retirement by Comptroller, 
    # Included - employer contributions from agencies/organizations/etc.
    
 #judges retirement OUT of fund
    # state borrowed money from pension funds to pay for core services during 2010 and 2011. 
    # used to fill budget gap and push problems to the future. 
    
    fund == "0319" ~ 4, # pension stabilization fund
    TRUE ~ 0) )

table(pension_totals$pension) 



pension_totals %>% 
  filter(pension != 0) %>%
  mutate(pension = as.factor(pension))%>%
  group_by(fy, pension) %>% 
  summarize(expenditure = sum(expenditure/1000000000, na.rm = TRUE)) %>%
  ggplot(aes(x=fy, y = expenditure, group=pension)) + 
  theme_classic()+
  geom_col(aes(fill = pension)) + 

  labs (title = "Pension expenditures", 
  caption = "1 = State contributions INTO pension funds.  
  2 = Employer Contributions.   
  3 = Purchase of Investments anomaly in 2010 and 2011. 
  4 = pension stabilization fund")+
    theme(legend.position = "bottom") +
  scale_y_continuous(labels = scales::dollar, name = "$ Billions")


```



```{r}
#| label: create-pension-variable


exp_temp <-  exp_temp |> 
arrange(fund) %>%
  mutate(pension = case_when( 
  
            (object=="1298" &  # Purchase of Investments, Normally excluded
       (fy==2010 | fy==2011) & 
       (fund=="0477" | fund=="0479" | fund=="0481")) ~ 3,  
       (object=="4431") ~ 1, # 4431 = easy to find pension payments INTO fund

    
     (object>"1159" & object<"1166") & fund != "0183" & fund != "0193"   ~ 2, 
    # objects 1159 to 1166 are all considered Retirement by Comptroller, 
    # Included - employer contributions from agencies/organizations/etc.
    
 #judges retirement OUT of fund
    # state borrowed money from pension funds to pay for core services during 2010 and 2011. 
    # used to fill budget gap and push problems to the future. 
    
    fund == "0319" ~ 4, # pension stabilization fund
    TRUE ~ 0) )

```

```{r}
#| label: fig-pensions-POB
#| fig-cap: "Pension Expenditures"


# special accounting of pension obligation bond (POB)-funded contributions to JRS, SERS, GARS, TRS 

exp_temp <- exp_temp |> 
  # change object for 2010 and 2011, retirement expenditures were bond proceeds and would have been excluded
  mutate(object = ifelse((pension > 0 & in_ff == "0"), "4431", object)) |> 
  # changes weird teacher & judge retirement system  pensions object to normal pension object 4431
  mutate(pension =  ifelse(pension > 0 & in_ff == "0", 6, pension)) |> # coded as 6 if it was supposed to be excluded. 
  mutate(in_ff = ifelse(pension > 0, "1", in_ff))



```

Can also be thought of past commitments vs current contributions.
Cost of past commitments in the form of Pension benefits paid out, 



**Current Employees vs Retired Employees**

Current Employees: 
- Group Insurance Benefits  

Retired Employees:
- Deferred Compensation  
- Medicare Retirees and Survivors of State of Illinois Employees Group Insurance Program (SEGIP)    
  - Part of Medicare   
  
  
  
```{r}

exp_temp |>
  filter(fy==2024) |>
  filter((appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400") ) |> 
  group_by(agency, agency_name) |> # separates CHIP from health and human services and saves it as Medicaid
  summarize(expenditure = sum(expenditure))


```
  
  
  
Drop all cash transfers between funds, statutory transfers, and purchases of investments from expenditure data.


```{r}
#| label: drop-transfers

transfers_drop <- exp_temp |> filter(
  agency == "799" | # statutory transfers
           object == "1993" |  # interfund cash transfers
           object == "1298") # purchase of investments
transfers_drop # items being dropped, 

# always check to make sure you aren't accidentally dropping something of interest.

exp_temp <- anti_join(exp_temp, transfers_drop)


```

### State employee healthcare costs

> Re-commented this out: Commented out line of code that seperates healthcare costs. This should keep healthcare costs in the agency, similar to the change that was made for pensions.

agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments


It looks like healthcare costs shift to "Other Departments" which includes: 

- GOMB (507)  
- Human Rights (442)  
- Illinois Power Agency (445)  
- Labor (452)   
- State Lottery (458)   
- Veteran's Affairs (497)    



```{r}
#| include: false

exp_temp |> filter(org_name == "BUREAU OF BENEFITS") |> 
  group_by(fy, agency) |>
  summarize(expenditure = sum(expenditure) ) |>
  pivot_wider(names_from = "fy", values_from = "expenditure")

exp_temp |> 
  filter(org_name == "BUREAU OF BENEFITS") |> 
  group_by(fy, agency) |>
  summarize(expenditure = sum(expenditure) ) |>
  ggplot() + 
  geom_line(aes(x=fy, y=expenditure, group= agency, color = agency)) + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title="Bureau of Benefits Expenditures")

exp_temp |> 
  filter(org_name == "BUREAU OF BENEFITS") |> 
  group_by(fy, object) |>
  summarize(expenditure = sum(expenditure) ) |>
  ggplot() + 
  geom_line(aes(x=fy, y=expenditure, group= object, color = object)) + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title="Bureau of Benefits Expenditures")

```


```{r}
#| label: eehc1


#if observation is a group insurance contribution, then the expenditure amount is set to $0 (essentially dropped from analysis)

# pretend eehc is named group_insurance_contribution or something like that
# eehc coded as zero implies that it is group insurance
# if eehc=0, then expenditures are coded as zero for group insurance to avoid double counting costs


exp_temp <- exp_temp |> 
  mutate(eehc = ifelse(
    # group insurance contributions for 1998-2005 and 2013-present
   fund == "0001" & (object == "1180" | object =="1900") & agency == "416" & appr_org=="20", 0, 1) )|> 
  
  mutate(eehc = ifelse(
    # group insurance contributions for 2006-2012
    fund == "0001" & object == "1180" & agency == "478" 
    & appr_org=="80", 0, eehc) )|>
    
   # group insurance contributions from road fund
  # coded with 1900 for some reason??
    mutate(eehc = ifelse(
      fund == "0011" & object == "1900" & 
        agency == "416" & appr_org=="20", 0, eehc) ) |>
  
  mutate(expenditure = ifelse(eehc=="0", 0, expenditure)) |>

  mutate(agency = case_when(
    ## turns specific items into State Employee Healthcare (agency=904)
      fund=="0907" & (agency=="416" & appr_org=="20") ~ "904",   # central management Bureau of benefits using health insurance reserve
      fund=="0907" & (agency=="478" & appr_org=="80") ~ "904",   # agency = 478: healthcare & family services using health insurance reserve - stopped using this in 2012
       TRUE ~ as.character(agency))) |>
  
  mutate(agency_name = ifelse(
    agency == "904", "STATE EMPLOYEE HEALTHCARE", as.character(agency_name)),
         in_ff = ifelse(agency == "904", 1, in_ff),
         group = ifelse(agency == "904", "904", as.character(agency)))  
# creates group variable

# Default group = agency number

healthcare_costs <- exp_temp |> filter(group == "904")

```



```{r}
#| label: transfers-to-local


exp_temp <- exp_temp |> mutate(
  agency = case_when(fund=="0515" & object=="4470" & type=="08" ~ "971", # income tax to local governments
                     fund=="0515" & object=="4491" & type=="08" & sequence=="00" ~ "971", # object is shared revenue payments
                     fund=="0802" & object=="4491" ~ "972", #pprt transfer
                     fund=="0515" & object=="4491" & type=="08" & sequence=="01" ~ "976", #gst to local
                     fund=="0627" & object=="4472"~ "976" , # public transportation fund but no observations exist
                     fund=="0648" & object=="4472" ~ "976", # downstate public transportation, but doesn't exist
                     fund=="0515" & object=="4470" & type=="00" ~ "976", # object 4470 is grants to local governments
                    object=="4491" & (fund=="0188"|fund=="0189") ~ "976",
                     fund=="0187" & object=="4470" ~ "976",
                     fund=="0186" & object=="4470" ~ "976",
                    object=="4491" & (fund=="0413"|fund=="0414"|fund=="0415")  ~ "975", #mft to local
                  fund == "0952"~ "975", # Added Sept 29 2022 AWM. Transportation Renewal MFT
                    TRUE ~ as.character(agency)),
  
  agency_name = case_when(agency == "971"~ "INCOME TAX 1/10 TO LOCAL",
                          agency == "972" ~ "PPRT TRANSFER TO LOCAL",
                          agency == "975" ~ "MFT TO LOCAL",
                          agency == "976" ~ "GST TO LOCAL",
                          TRUE~as.character(agency_name)),
  group = ifelse(agency>"970" & agency < "977", as.character(agency), as.character(group)))

```


### Local Transfers

```{r}
#| label: fig-drop-local-transfers
#| fig-cap: "Drop Transfers from State to Local Governments"

transfers_long <- exp_temp |> 
  filter((group == "971" |group == "972" | group == "975" | group == "976")) 
#  fund == "0325")

transfers_long |> 
  group_by(agency_name, group, fy) |> 
  summarize(expenditure = sum(expenditure, na.rm=TRUE) )|> 
  ggplot() + 
  geom_line(aes(x=fy, y = expenditure, color=agency_name)) + 
  alea_theme() + 
  scale_x_continuous(expand = c(0,0), limits = c(1998, current_year+.5), breaks = c(1998, 2005, 2010, 2015, 2020, current_year)) +
  
  labs(title = "Transfers to Local Governments", caption = "Data Source: Illinois Office of the Comptroller")

```




```{r}
transfers_long <- exp_temp |> 
  filter(group == "971" |group == "972" | group == "975" | group == "976")


transfers <- transfers_long |>
  group_by(fy, group ) |>
  summarize(sum_expenditure = sum(expenditure)/1000000) |>
  pivot_wider(names_from = "group", values_from = "sum_expenditure", names_prefix = "exp_" )

exp_temp <- anti_join(exp_temp, transfers_long)


dropped_inff_0 <- exp_temp |> filter(in_ff == 0)

exp_temp <- exp_temp |> filter(in_ff == 1) # drops in_ff = 0 funds AFTER dealing with net-revenue above

```


### Debt


```{r}
#| label: debt-service

debt_drop <- exp_temp |> 
  filter(object == "8841" |  object == "8811")  
# escrow  OR  principle

#debt_drop |> group_by(fy) |> summarize(sum = sum(expenditure)) |> arrange(-fy)


debt_keep <- exp_temp |> 
  filter(fund != "0455" & (object == "8813" | object == "8800" )) 
# examine the debt costs we want to include

#debt_keep |> group_by(fy) |> summarize(sum = sum(expenditure)) |> arrange(-fy) 


exp_temp <- anti_join(exp_temp, debt_drop) 
exp_temp <- anti_join(exp_temp, debt_keep)

debt_keep <- debt_keep |>
  mutate(
    agency = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(agency)),
    group = ifelse(fund != "0455" & (object == "8813" | object == "8800"), "903", as.character(group)),
    in_ff = ifelse(group == "903", 1, as.character(in_ff)))

debt_keep_yearly <- debt_keep |> 
  group_by(fy, group) |> 
  summarize(debt_cost = sum(expenditure,na.rm=TRUE)/1000000) |> 
  select(-group)

```

```{r}


tollway_exp <- exp_temp |> filter(fund == "0455") |> group_by(fy) |> summarize(expenditure = sum(expenditure))
                                                                                                    #tollway_exp |> ggplot() + geom_line(aes(x=fy, y=expenditure)) + labs(title = "Fund 0455 from Expenditure: All Tollway Expenditures", caption = "Data from IOC Expenditure Files. Fund 0455 is the IL State Tollway")


# all tollway revenues, not just bond proceeds
alltollway<-rev_temp |> filter(fund == "0455" & source != "0571") |> group_by(fy) |> summarize(sum = sum(receipts, na.rm = TRUE))


# tollway bond proceeds
tollway_bondproc <- rev_temp |> filter(fund == "0455" & source == "0571" ) |> group_by(fy) |> summarize(sum = sum(receipts, na.rm = TRUE))

#alltollway |>  ggplot() + geom_line(aes(x=fy, y=sum)) + labs(title = "Fund 0455 - All Tollway Revenue", caption = "Data from IOC Revenue Files. Fund 0455 is the IL State Tollway Revenue") 

#tollway_bondproc |> ggplot() + geom_line(aes(x=fy, y=sum)) + labs(title = "Fund 0455 - Tollway Revenue: Tollway Bond Proceeds", caption = "Data from IOC Revenue Files. Fund 0455 is the IL State Tollway Revenue")
  
#ggplot() + geom_line(data=tollway_bondproc, aes(x=fy, y=sum)) + labs(title = "Fund 0455 - Tollway Revenue: Tollway Bond Proceeds", caption = "Data from IOC Revenue Files. Fund 0455 is the IL State Tollway Revenue")

#tollwaydebt |> ggplot() + geom_line(aes(x=fy, y=sum)) + labs(title = "Tollway Debt Service", caption = "Debt service includes principal and interest for the Illinois Tollway. Object = 8800 and fund = 0455")


#tollway debt principal and interest
tollwaydebt <- exp_temp |>filter(object == "8800" & fund == "0455") |> group_by(fy) |> summarize(sum=sum(expenditure)) 

# Tollway agency expenditures = SAME as filtering by fund == 0455
#tollway<-exp_temp |> filter(agency == "557")
#exp_temp |> filter(agency == "557") |> group_by(fy) |> summarize(sum = sum(expenditure)) |> arrange(-fy)

# contributions and benefits paid comparison
ggplot()+
  scale_x_continuous(expand = c(0,0), limits = c(1998, current_year+.5), breaks = c(1998, 2005, 2010, 2015, 2020, current_year)) +
    geom_line(data=tollway_bondproc, aes(x=fy, y=sum, color='Bond Proceeds')) +
  geom_line(data= tollwaydebt, aes(x=fy, y = sum, color = 'Debt Service'))+ 
  geom_line(data= tollway_exp, aes(x=fy, y = expenditure, color = 'Tollway Expenditures'))+ 
  geom_line(data= alltollway, aes(x=fy, y = sum, color = "Tollway Revenue"))+ 
  scale_color_manual(values = c(
    'Bond Proceeds' = 'darkblue',
    'Debt Service' = 'red',
    'Tollway Expenditures' = 'orange',
    'Tollway Revenue' = 'light green')) +
  labs(title="Tollway bond procreeds, debt service, revenue, and expenditures.", 
       caption = "Tollway revenue + bond proceeds should be roughly equal to tollway expenditures + debt service.", 
       y = "Dollars")
```


### Add Other Fiscal Future group codes


> Commented out line that creates Other departments from a few agencies. Keeping them seperate to see where healthcare costs go. 

agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments

<!--- if for the IGPA datahub visualizations, combine smallest categories to "Other Departments". For the download data link, keep them separate. At least that is the current request as of March 14 2026   --->

```{r}
#| label: group-codes


exp_temp <- exp_temp |>
  #mutate(agency = as.numeric(agency) ) |>
  # arrange(agency)|>
  mutate(
    group = case_when(

             pension > 0  ~ "901", # pensions,
       
      agency>"100"& agency<"200" ~ "910", # legislative
      
      agency == "528"  | (agency>"200" & agency<"300") ~ "920", # judicial 
      
      ######################################################
  
  
      # Not used if we are keeping separate pension costs!!    
      # ## New CODE: April 23rd, 2025:
      # agency == "593" ~ "959", #  TRS becomes part of K-12 costs
      # agency == "594" ~ "959",   # TRS
      # agency == "589" ~ "589", # SERS becomes part of "Other Agencies"
      # agency == "693" ~ "960", # SURS becomes part of group 960
      # agency == "275" ~ "920",  # JRS becomes part of group 920
      # agency == "131" ~  "910", # GARS becomes part of Group 910
      ######################################################
      
      (agency>"309" & agency<"400") ~ "930",    # elected officers: Governor, lt gov, attorney general, sec. of state, comptroller, treasurer
      
      agency == "586" ~ "959", # create new K-12 group

      agency=="402" | agency=="418" | agency=="478" | agency=="444" | agency=="482" ~ as.character(agency), # aging, CFS, HFS, human services, public health
      T ~ as.character(group))
    ) |>      

  
  mutate(group = case_when(
        # Not used if we are not separating pension costs!!
       pension > 0  ~ "901", # pensions,
       
    agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400") ~ "945", # separates CHIP from health and human services and saves it as Medicaid
    
    agency == "586" & fund == "0355" ~ "945",  # 586 (Board of Edu) has special education which is part of medicaid
    
    # OLD CODE: agency == "586" & appr_org == "18" ~ "945", # Spec. Edu Medicaid Matching
    
    agency=="425" | agency=="466" | agency=="546" | agency=="569" | agency=="578" | agency=="583" | agency=="591" | agency=="592" | agency=="493" | agency=="588" ~ "941", # public safety & Corrections
    
    agency=="420" | agency=="494" |  agency=="406" | agency=="557" ~ as.character(agency), # econ devt & infra, tollway
    
    agency=="511" | agency=="554" | agency=="574" | agency=="598" ~ "946",  # Capital improvement
    
    agency=="422" | agency=="532" ~ as.character(agency), # environment & nat. resources
    
    agency=="440" | agency=="446" | agency=="524" | agency=="563"  ~ "944", # business regulation
    
    agency=="492" ~ "492", # revenue
    
    agency == "416" ~ "416", # central management services
    agency=="448" & fy > 2016 ~ "416", #add DoIT to central management 
    
    T ~ as.character(group))) |>
  
  
  mutate(group = case_when(
        # Not used if we are not separating pension costs!!
       pension > 0  ~ "901", # pensions
       
    # agency=="684" | agency=="691"  ~ as.character(agency), # moved under higher education in next line. 11/28/2022 AWM
    
    agency=="692" | agency == "693" | agency=="695" | agency == "684" |agency == "691" | (agency>"599" & agency<"677") ~ "960", # higher education
    
    agency=="427"  ~ as.character(agency), # employment security
    
############################ 
# Leaving these agencies as their own agency number for now. Had been coded to "Other departments" Group 948
# - GOMB (507)  
# - Human Rights (442)  
# - Illinois Power Agency (445)  
# - Labor (452)   
# - State Lottery (458)   
# - Veteran's Affairs (497) 

       agency=="507" | agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ as.character(agency), # Were included within "other departments"
    
#    agency=="507"|  agency=="442" | agency=="445" | agency=="452" |agency=="458" | agency=="497" ~ "948", # other departments

###########################################
    

# other boards & Commissions
    agency=="503" | agency=="509" | agency=="510" | agency=="565" |agency=="517" | agency=="525" | agency=="526" | agency=="529" | agency=="537" | agency=="541" | agency=="542" | agency=="548" |  agency=="555" | agency=="558" | agency=="559" | agency=="562" | agency=="564" | agency=="568" | agency=="579" | agency=="580" | agency=="587" | agency=="590" | agency=="527" | agency=="585" | agency=="567" | agency=="571" | agency=="575" | agency=="540" | agency=="576" | agency=="564" | agency=="534" | agency=="520" | agency=="506" | agency == "533" ~ "949", 
    
# Other Departments
  #   Before pensions were included back with the original agency that spent the money, remaining non-pension expenditures from agencies that deal with pensions were included with Other Departments 
    agency=="131" |
    agency=="275" | #JRS
    agency=="589" | #SERS
    agency=="593"|  # TRS
    agency=="594"| # Also TRS
    agency=="693"   #SURS
   ~ "948",
    
    T ~ as.character(group))) |>
  
  mutate(group_name = 
           case_when(
             group == "416" ~ "Central Management",
             group == "442" ~ "Human Rights",
             group == "445" ~ "IL Power Agency",
             group == "452" ~ "Labor",
             group == "458" ~ "State Lottery",
             group == "589" ~ "SERS",
             group == "478" ~ "Healthcare and Family Services",
             group == "482" ~ "Public Health",
             group == "901" ~ "State Pension Contributions", ## Split up into GARS, SERS, etc. now
             group == "903" ~ "Debt Service",
             group == "910" ~ "Legistlative"  ,
             group == "920" ~ "Judicial" ,
             group == "930" ~ "Elected Officers" , 
             group == "941" ~ "Public Safet" ,
             group == "942" ~ "Econ Development & Infrastructure" ,
             group == "943" ~ "Central Services",
             group == "944" ~ "Business & Professional Regulation" ,
             group == "945" ~ "Medicaid" ,
             group == "946" ~ "Capital Improvement" , 
             group == "948" ~ "Other Departments" ,
             group == "949" ~ "Other Boards & Commissions" ,
             group == "959" ~ "K-12 Education" ,
             group == "960" ~ "University Education" ,
             group == agency ~ as.character(agency_name),
             TRUE ~ "Check name"),
         year = fy)


exp_temp <- exp_temp |>
    mutate(fund_cat_name =
           case_when(
             fund_cat_name == "General Fund" ~ "General Funds",
             fund_cat_name == "REVOLVING FUNDS" ~ "Revolving Funds",
             T ~ fund_cat_name
           ),
         federal_funded = case_when(
           fund_cat_name == "Federal Trust Funds" ~ "Federal Funds",
           group_name == "MEDICAID" & fund_cat_name == "General Funds" ~ "Federal Funds",
          T ~ "State Funds"
           
         ))


exp_temp |> filter(group_name == "Check name")


```

::: callout-important
All expenditures recoded but not aggregated: Allows for inspection of individual expenditures within larger categories. This stage of the data is extremely useful for investigating how individual items have been coded before they are aggregated into larger categories.
:::

## Modify Revenue data

```{r}
#| label: rev-recode-agencies

# recodes old agency numbers to consistent agency number
rev_temp <- rev_temp |> 
  mutate(agency = case_when(
    (agency=="438"| agency=="475" |agency == "505") ~ "440",
    # financial institution &  professional regulation &
     # banks and real estate  --> coded as  financial and professional reg
    agency == "473" ~ "588", # nuclear safety moved into IEMA
    (agency =="531" | agency =="577") ~ "532", # coded as EPA
    (agency =="556" | agency == "538") ~ "406", # coded as agriculture
    agency == "560" ~ "592", # IL finance authority (fire trucks and agriculture stuff)to state fire marshal
    agency == "570" & fund == "0011" ~ "494",   # city of Chicago road fund to transportation
    TRUE ~ (as.character(agency)))) |>
  mutate(fund_cat_name =
           case_when(
             fund_cat_name == "General Fund" ~ "General Funds",
             fund_cat_name == "REVOLVING FUNDS" ~ "Revolving Funds",
             T ~ fund_cat_name
           ))
```

#### Federal Transfers

```{r}
#| label: fig-create-rev-federal-transfers

#rev_temp <- rev_temp |> filter(in_ff==1)

rev_temp <- rev_temp |> 
  mutate(
    rev_type = ifelse(rev_type=="57" & agency=="478" & (source=="0618"|source=="2364"|source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692"), "58", rev_type),
    rev_type_name = ifelse(rev_type=="58", "Federal Medicaid Reimbursements", rev_type_name),
    rev_type = ifelse(rev_type=="57" & agency=="494", "59", rev_type),
    rev_type_name = ifelse(rev_type=="59", "Federal Transportation", rev_type_name),
    rev_type_name = ifelse(rev_type=="57", "Federal - Other", rev_type_name),
    rev_type = ifelse(rev_type=="6", "06", rev_type),
    rev_type = ifelse(rev_type=="9", "09", rev_type)) 

rev_temp |> 
  filter(rev_type == "58" | rev_type == "59" | rev_type == "57") |> 
  group_by(fy, rev_type, rev_type_name) |> 
  summarise(receipts = sum(receipts, na.rm = TRUE)/1000000) |> 
  ggplot() +
  geom_line(aes(x=fy, y=receipts,color=rev_type_name)) +
  scale_y_continuous(labels = comma)+
  labs(title = "Federal to State Transfers", 
       y = "Millions of Dollars", x = "") + 
  theme(legend.position = "bottom", legend.title = element_blank()  )

```

```{r}
#| label: coding-stimuluspackage-dollars

rev_temp <- rev_temp |> mutate(covid_dollars = ifelse(source_name_AWM == "FEDERAL STIMULUS PACKAGE",1,0))

rev_temp |> filter(source_name_AWM == "FEDERAL STIMULUS PACKAGE") |>
  group_by(fy) |> summarize(Received = sum(receipts))
  
```

### Medicaid Revenue vs Medicaid Expenditures


object 4900 is Lump Sums, object 4400 is Awards & Grants.

```{r}

medicaid_cost_total <- exp_temp |> 
  filter(agency=="478" & (appr_org=="01" | appr_org == "65" | appr_org=="88") & (object=="4900" | object=="4400")) |> 
  group_by(fy) |> 
  summarize(sum=sum(expenditure, na.rm=TRUE))

medicaid_cost <- exp_temp |> 
  filter(#agency=="478" & 
          # (appr_org=="01" | appr_org == "65" | appr_org=="88") & 
           (object=="4900" | object=="4400")) |> 
  group_by(fy, agency) |> 
  summarize(sum=sum(expenditure, na.rm=TRUE))

ggplot()+
  geom_line(data=medicaid_cost_total, aes(x=fy, y = sum, color = "Expenditures"), lwd = 1) +
    geom_line(data=medicaid_cost, aes(x=fy, y = sum, color = agency)) + 

  scale_x_continuous(n.breaks = 6) +
  labs(title = "Medicaid expenditures", 
       caption = "Medicaid expenditures include funds provided to medical providers.", 
       color = element_blank()
       )
```


### Health Insurance Premiums from Employees

Insurance premiums for employees is coded below but it is NOT used in the fiscal futures model. Employee and employer premiums are considered rev_51 and dropped from analysis in later step.

-   0120 = ins prem-option life
-   0120 = ins prem-optional life/univ
-   0347 = optional health - HMO
-   0348 = optional health - dental
-   0349 = optional health - univ/local SI
-   0350 = optional health - univ/local
-   0351 = optional health - retirement
-   0352 = optional health - retirement SI
-   0353 = optional health - retire/dental
-   0354 = optional health - retirement hmo
-   2199-2209 = various HMOs, dental, health plans from Health Insurance Reserve (fund)


```{r}
#| label: insurance-premiums

#collect optional insurance premiums to fund 0907 for use in eehc expenditure  
rev_temp <- rev_temp |> 
  mutate(
    employee_premiums = ifelse(fund=="0907" & (source=="0120"| source=="0121"| (source>"0345" & source<"0357")|(source>"2199" & source<"2209")), 1, 0),
    
    # adds more rev_type codes
    rev_type = case_when(
      fund =="0427" ~ "12", # pub utility tax
      fund == "0742" | fund == "0473" ~ "24", # insurance and fees
      fund == "0976" ~ "36",# receipts from rev producing
      fund == "0392" |fund == "0723" ~ "39", # licenses and fees
      fund == "0656" ~ "78", #all other rev sources
      TRUE ~ as.character(rev_type)))
# if not mentioned, then rev_type as it was
```


```{r}
#| label: insurance-premiums-long


# # optional insurance premiums = employee insurance premiums

# emp_premium <- rev_temp |>
#   group_by(fy, employee_premiums) |>
#   summarize(employee_premiums_sum = sum(receipts)/1000000) |>
#   filter(employee_premiums == 1) |>
#   rename(year = fy) |> 
#   select(-employee_premiums)

emp_premium_long <- rev_temp |>  filter(employee_premiums == 1)
emp_premium_long

# drops employee premiums from revenue
# rev_temp <- rev_temp |> filter(employee_premiums != 1)
# should be dropped in next step since rev_type = 51


```

*Note: In FY21, employee premiums were subtracted from state healthcare costs on the expenditure side to calculate a "Net Healthcare Cost" but that methodology has been discontinued. Totals were practically unchanged: revenue from employee premiums is also very small.*

### Transfers in and Out:

Funds that hold and disperse local taxes or fees are dropped from the analysis. Then other excluded revenue types are also dropped.

Drops Blank, Student Fees, Retirement contributions, proceeds/investments, bond issue proceeds, interagency receipts, cook IGT, Prior year refunds:

> Clean up code and annotations in chunk below:

```{r}

rev_temp <- rev_temp |> 
  filter(in_ff == 1) |> 
  mutate(local = ifelse(is.na(local), 0, local)) |> # drops all revenue observations that were coded as "local == 1"
  filter(local != 1)

# 1175 doesnt exist?
in_from_out <- c("0847", "0867", "1175", "1176", "1177", "1178", "1181", "1182", "1582", "1592", "1745", "1982", "2174", "2264")

# what does this actually include:
# all are items with rev_type = 75 originally. 
in_out_df <- rev_temp |>
  mutate(infromout = ifelse(source %in% in_from_out, 1, 0)) |>
  filter(infromout == 1)

rev_temp <- rev_temp |> 
   mutate(rev_type_new = ifelse(source %in% in_from_out, "76", rev_type))
# if source contains any of the codes in in_from_out, code them as 76 (all other rev).
# I end up excluding rev_76 in later steps

```


```{r}
#| label: droprevtypes


# revenue types to drop
drop_type <- c("32", "45", "51", 
               "66", "72", "75", "76", "79", "98", "99")

# drops Blank, Student Fees, Retirement contributions, proceeds/investments,
# bond issue proceeds, interagency receipts, cook IGT, Prior year refunds.


rev_temp <- rev_temp |> filter(!rev_type_new %in% drop_type)
# keep observations that do not have a revenue type mentioned in drop_type

table(rev_temp$rev_type_new)

```

## Pivoting and Merging


### Revenues

<!--- If there are NA rev types, it will cause the code to  break when binding dataframes together a few chunks lower  --->

```{r}
#| label: code--check
#| include: false

rev_temp |> filter(is.na(rev_type))

```

```{r}
#| label: make-ffrev-table

ff_rev <- rev_temp |> 
  group_by(rev_type_new, fy) |> 
  summarize(sum_receipts = sum(receipts, na.rm=TRUE)/1000000 ) |>
  pivot_wider(names_from = "rev_type_new", values_from = "sum_receipts", names_prefix = "rev_")

 
ff_rev <- mutate_all(ff_rev, replace_na, 0)

```


```{r}
#| eval: false
#| include: false

# DOCUMENTATION PURPOSES ONLY: If refunds on the expenditure side are subtracted from the revenue side, it was done like this. The refunds were identified in the "Tax Refunds" section of the code. 

# OLD way of doing refunds ##
# ff_rev <- ff_rev |>
#   mutate(rev_02 = rev_02 - ref_02,
#          rev_03 = rev_03 - ref_03,
#          rev_06 = rev_06 - ref_06,
#          rev_09 = rev_09 - ref_09,
#          rev_21 = rev_21 - ref_21,
#          rev_24 = rev_24 - ref_24,
#          rev_35 = rev_35 - ref_35
# 
#       #   rev_78new = rev_78 #+ pension_amt #+ eehc
#          ) |> 
#   select(-c(ref_02:ref_35, rev_99, rev_NA, rev_76
#             #, ref_CHECK#, pension_amt , rev_76,
#           #  , eehc
#             ))
# 
# ff_rev



#noproblem <- c(0)  # if ref_CHECK = $0, then there is no problem. :) 
# 
# if((sum(ff_rev$ref_CHECK) == 0 )){
# 
# ff_rev <- ff_rev |>
#   
#   mutate(rev_02 = rev_02 - ref_02,
#          rev_03 = rev_03 - ref_03,
#          rev_06 = rev_06 - ref_06,
#          rev_09 = rev_09 - ref_09,
#          rev_21 = rev_21 - ref_21,
#          rev_24 = rev_24 - ref_24,
#          rev_35 = rev_35 - ref_35
#          ) |> 
#   select(-c(ref_02:ref_35, rev_99, rev_76, ref_CHECK )) 
# }else{"You have a problem! Check what revenue items did not have rev codes (causing it to be coded as rev_NA) or the check if there were refunds that were not assigned revenue codes (tax_refunds_long objects)"}

```


```{r}
#| label: tbl-labeled-final-ffrev-table
#| tbl-cap: "Aggregated Revenue Categories ($ Millions), with old labels. Not actually used for anything other than to have output that was easiy comparable to old STATA output from past years."
#| include: false

# Since I already pivot_wider()ed the table in the previous code chunk, I now change each column's name by using rename() to set new variable names. Ideally the final dataframe would have both the variable name and the variable label but I have not done that yet.


aggregate_rev_labels <- ff_rev |>
  rename("INDIVIDUAL INCOME TAXES, gross of local, net of refunds" = rev_02,
         "CORPORATE INCOME TAXES, gross of PPRT, net of refunds" = rev_03,
         "SALES TAXES, gross of local share" = rev_06 ,
         "MOTOR FUEL TAX, gross of local share, net of refunds" = rev_09 ,
         "PUBLIC UTILITY TAXES, gross of PPRT" = rev_12,
         "CIGARETTE TAXES" = rev_15 ,
         "LIQUOR GALLONAGE TAXES" = rev_18,
         "INHERITANCE TAX" = rev_21,
         "INSURANCE TAXESFEES&LICENSES, net of refunds" = rev_24 ,
         "CORP FRANCHISE TAXES & FEES" = rev_27,
        "HORSE RACING TAXES & FEES" = rev_30,  # in Other
         "MEDICAL PROVIDER ASSESSMENTS" = rev_31 ,
         # "GARNISHMENT-LEVIES " = rev_32 , # dropped
         "LOTTERY RECEIPTS" = rev_33 ,
         "OTHER TAXES" = rev_35,
         "RECEIPTS FROM REVENUE PRODUCNG" = rev_36, 
         "LICENSES, FEES & REGISTRATIONS" = rev_39 ,
         "MOTOR VEHICLE AND OPERATORS" = rev_42 ,
         #  "STUDENT FEES-UNIVERSITIES" = rev_45,   # dropped
         "RIVERBOAT WAGERING TAXES" = rev_48 ,
         # "RETIREMENT CONTRIBUTIONS " = rev_51, # dropped
         "GIFTS AND BEQUESTS" = rev_54, 
         "FEDERAL OTHER" = rev_57 ,
         "FEDERAL MEDICAID" = rev_58, 
         "FEDERAL TRANSPORTATION" = rev_59 ,
         "OTHER GRANTS AND CONTRACTS" = rev_60, #other
        "INVESTMENT INCOME" = rev_63, # other
         # "PROCEEDS,INVESTMENT MATURITIES" = rev_66 , #dropped
         # "BOND ISSUE PROCEEDS" = rev_72,  #dropped
         # "INTER-AGENCY RECEIPTS" = rev_75,  #dropped
       # "TRANSFER IN FROM OUT FUNDS" = rev_76,  # dropped
         "ALL OTHER SOURCES" = rev_78,
         # "COOK COUNTY IGT" = rev_79, #dropped
         # "PRIOR YEAR REFUNDS" = rev_98 #dropped
  ) 

aggregate_rev_labels |> mutate_all(round, digits = 0)

```

### Expenditures

Create exp_970 for all local government transfers (exp_971 + exp_972 + exp_975 + exp_976).

```{r}
#| label: tbl-ffexp-notlabeled
#| tbl-cap: "Pivoted Expenditure Categories ($ Millions). Intermediate step. Not actually used for anything other than having output similar to past STATA output."

ff_exp <- exp_temp |> 
  group_by(fy, group) |> 
  summarize(sum_expenditures = sum(expenditure, na.rm=TRUE)/1000000 ) |>
  pivot_wider(names_from = "group", values_from = "sum_expenditures", names_prefix = "exp_")|>
  
    left_join(debt_keep_yearly) |>
  rename(exp_903 = debt_cost) |>

 #  join local transfers and create exp_970
  left_join(transfers) |>
  mutate(exp_970 = rowSums(across(any_of(c("exp_971", "exp_972", "exp_975", "exp_976"))), na.rm = TRUE))

 ff_exp<- ff_exp |> 
select(-any_of(paste0("exp_", 971:976)))
 # drop unwanted columns that are already included in exp_970

ff_exp # not labeled

```


# All Funds vs General Funds

**Work in progress!**

```{r}
rev_fund_cats <- rev_temp |> 
  mutate(fund_cat_name = ifelse(fund_cat_name == "REVOLVING FUNDS", "Revolving Funds", fund_cat_name) ) |>
           filter(in_ff == 1) |> 
  group_by(fy, fund_cat_name) |> 
  summarize(Revenue = sum(receipts, na.rm=TRUE))

rev_fund_cats |> filter(fy == 2025)

exp_fund_cats <- exp_temp |>   mutate(fund_cat_name = ifelse(fund_cat_name == "REVOLVING FUNDS", "Revolving Funds", fund_cat_name) ) |>
filter(in_ff == 1) |> group_by(fy, fund_cat_name) |> summarize(Expenditures = sum(expenditure, na.rm=TRUE))

exp_fund_cats |> filter(fy == 2025)


ggplot() +
  geom_line(data = rev_fund_cats, aes(x=fy, y = Revenue, color = fund_cat_name)) + 
    geom_line(data = exp_fund_cats, aes(x=fy, y = Expenditures, color = fund_cat_name), lty=2)  + 
  scale_y_billions(name = "$")

```



```{r}
rev_fund_cats <- rev_temp |>   
  mutate(fund_cat_name = ifelse(fund_cat_name == "REVOLVING FUNDS", "Revolving Funds", fund_cat_name) ) |>
filter(in_ff == 1 & fund_cat_name == "General Funds") |> group_by(fy, fund_cat_name) |> summarize(Revenue = sum(receipts, na.rm=TRUE))
rev_fund_cats |> filter(fy == 2025)

exp_fund_cats <- exp_temp |> 
    mutate(fund_cat_name = ifelse(fund_cat_name == "REVOLVING FUNDS", "Revolving Funds", fund_cat_name) ) |>
filter(in_ff == 1 & fund_cat_name == "General Funds") |> group_by(fy, fund_cat_name) |> summarize(Expenditures = sum(expenditure, na.rm=TRUE))
exp_fund_cats |> filter(fy == 2025)


ggplot() +
  geom_line(data = rev_fund_cats, aes(x=fy, y = Revenue)) + 
  geom_line(data = exp_fund_cats, aes(x=fy, y = Expenditures), lty=2)  + 
  scale_y_billions(name = "$") +
  labs(caption = "Dashed line is General Fund Expenditures",
       title = "General Fund Fiscal Gap",
       subtitle = "Pretend there is a line showing there is a difference between the two lines.",
       legend = "",
       x = element_blank())

```

# Graphs and Tables

Create total revenues and total expenditures only:

-   after aggregating expenditures and revenues, pivoting wider, then I want to drop the columns that I no longer want and then pivot_longer(). After pivoting_longer() and creating `rev_long` and `exp_long`, expenditures and revenues are in the same format and can be combined together for the totals and gap each year.


```{r}
#| label: tbl-rev-and-exp-totals-long
#| tbl-cap: "Long Version of Data that has Revenue and Expenditures in One Dataframe. Creates `expenditures_recoded_long_pensionchange_FY`, `revenues_recoded_long_pensionchange_FY` and `aggregated_totals_pensionchange` which are exported as CSVs."
#| tbl-cap-location: top

rev_long <- pivot_longer(ff_rev, rev_02:rev_78, names_to = c("type","Category"), values_to = "Dollars", names_sep = "_") |> 
  rename(Year = fy) |>
  mutate(Category_name = case_when(
    Category == "02" ~ "INDIVIDUAL INCOME TAXES" ,
    Category == "03" ~ "CORPORATE INCOME TAXES" ,
    Category == "06" ~ "SALES TAXES" ,
    Category == "09" ~ "MOTOR FUEL TAX" ,
    Category == "12" ~ "PUBLIC UTILITY TAXES" ,
    Category == "15" ~ "CIGARETTE TAXES" ,
    Category == "18" ~ "LIQUOR GALLONAGE TAXES" ,
    Category == "21" ~ "INHERITANCE TAX" ,
    Category == "24" ~ "INSURANCE TAXES, FEES & LICENSES" ,
    Category == "27" ~ "CORP FRANCHISE TAXES & FEES" ,
    Category == "30" ~ "HORSE RACING TAXES & FEES", 
    Category == "31" ~ "MEDICAL PROVIDER ASSESSMENTS" ,
    Category == "32" ~ "GARNISHMENT-LEVIES" , # dropped
    Category == "33" ~  "LOTTERY RECEIPTS" ,
    Category == "35" ~  "OTHER TAXES" ,
    Category == "36" ~  "RECEIPTS FROM REVENUE PRODUCING", 
    Category == "39" ~  "LICENSES, FEES & REGISTRATIONS" ,
    Category == "42" ~  "MOTOR VEHICLE AND OPERATORS" ,
    Category == "45" ~  "STUDENT FEES-UNIVERSITIES",   # dropped
    Category == "48" ~  "RIVERBOAT WAGERING TAXES" ,
    Category == "51" ~  "RETIREMENT CONTRIBUTIONS" , # dropped
    Category == "54" ~ "GIFTS AND BEQUESTS", 
    Category == "57" ~  "FEDERAL OTHER" ,
    Category == "58" ~  "FEDERAL MEDICAID", 
    Category == "59" ~  "FEDERAL TRANSPORTATION" ,
    Category == "60" ~  "OTHER GRANTS AND CONTRACTS", 
    Category == "63" ~  "INVESTMENT INCOME", 
    Category == "66" ~ "PROCEEDS, INVESTMENT MATURITIES" , #dropped
    Category == "72" ~ "BOND ISSUE PROCEEDS",  #dropped
    Category == "75" ~  "INTER-AGENCY RECEIPTS ",  #dropped
    Category == "76" ~  "TRANSFER IN FROM OUT FUNDS",  
    Category == "78" ~  "ALL OTHER SOURCES" ,
    Category == "79" ~   "COOK COUNTY IGT", #dropped
    Category == "98" ~  "PRIOR YEAR REFUNDS", #dropped
    T ~ "Check Me!"
    
  ) )|> 
  mutate(Category_name = str_to_title(Category_name))


exp_long <- pivot_longer(ff_exp, exp_402:exp_970 , names_to = c("type", "Category"), values_to = "Dollars", names_sep = "_") |> 
  rename(Year = fy ) |> 
  mutate(Category_name = 
           case_when(
             
             Category == "131" ~ "GARS",    # should be in Legislative Group already
             Category == "275" ~ "JRS",     # should be in Judicial Category already
             Category == "402" ~ "Aging",
             Category == "406" ~ "Agriculture",   # agriculture
             Category == "416" ~ "Central Management", ## contains DoIT also
             Category == "418" ~ "Children & Family Services", 
             Category == "420" ~ "Commerce & Economic Opportunity",
             Category == "422" ~ "Natural Resources" ,
             Category == "426" ~ "Corrections",
             Category == "427" ~ "Employment Security" ,
             
             Category == "442" ~ "Human Rights" ,  # sometimes included in "Other Departments" when trying to have fewer expenditure categories
             
             Category == "444" ~ "Human Services" ,  
             Category == "445" ~ "IL Power Agency" ,    # IL Power Agency
             Category == "448" ~ "Innovation & Technology",   # should be in Central Management already
             
             Category == "452" ~ "Labor" ,   # Sometimes included in "Other Departments when trying to have fewer categories
             
             Category == "458" ~ "State Lottery" ,   # State Lottery is sometimes included as "Other Departments when trying to have fewer expenditure categories
             
             Category == "478" ~ "Family Services (net Medicaid)",
             Category == "482" ~ "Public Health", 
             Category == "492" ~ "Revenue", 
             Category == "493" ~ "Teacher Retirement System (TRS)",  # Should be included in K-12 already
             Category == "494" ~ "Transportation" ,
             Category == "507" ~ "GOMB",  # GOMB     # GOMB is sometimes included as "Other Departments when trying to have fewer expenditure categories
             Category == "497" ~ "Veterans' Affairs",  # Veterans' Affairs  is sometimes included as "Other Departments when trying to have fewer expenditure categories
             Category == "532" ~ "Environmental Protection Agency" ,
             Category == "557" ~ "IL State Tollway" ,
             Category == "589" ~ "State Emp. Retirement System (SERS)",
             Category == "693" ~ "SURS",                             # should be in Higher Education already
             Category == "901" ~ "Pension Expenditure",
             Category == "903" ~ "Debt Service",
             Category == "904" ~ "State Employee Healthcare",
             Category == "910" ~ "Legislative"  ,
             Category == "920" ~ "Judicial" ,
             Category == "930" ~ "Elected Officers" , 
             Category == "941" ~ "Public Safety" ,
             Category == "943" ~ "Central Services",
             Category == "944" ~ "Business & Professional Regulation" ,
             Category == "945" ~ "Medicaid" ,
             Category == "946" ~ "Capital Improvements" , 
             Category == "948" ~ "Other Departments" ,   # Used when pre-grouping small agencies to group = 948. 
             Category == "949" ~ "Other Boards & Commissions" ,
             Category == "959" ~ "K-12 Education" ,
             Category == "960" ~ "University Education",
             Category == "970" ~ "Local Govt Transfers",
             T ~ "CHECK ME!")
           )   |>
  dplyr::mutate(Dollars = ifelse(is.na(Dollars), 0, Dollars))


# combine revenue and expenditures into one data frame
aggregated_totals_long <- rbind(rev_long, exp_long)

# 
# if(any(aggregated_totals_long$Category_name == "CHECK ME!" & aggregated_totals_long$Dollars>0)){
#   
#   aggregated_totals_long <- aggregated_totals_long
#    
# }else{
#   aggregated_totals_long <- aggregated_totals_long |> 
#     filter(!Category_name %in% c("CHECK ME!", "Check Me!")) |>
#     mutate(Dollars = ifelse(is.na(Dollars), 0, Dollars))
# }
# 
# if(!any(aggregated_totals_long$Category_name == "CHECK ME!" & 
#         (aggregated_totals_long$Dollars > 0 | is.na(aggregated_totals_long$Dollars)) ) ) {
# 
#   aggregated_totals_long <- aggregated_totals_long |>
#     filter(!Category_name %in% c("CHECK ME!", "Check Me!")) |>
#     mutate(Dollars = ifelse(is.na(Dollars), 0, Dollars))
# }


if (any(aggregated_totals_long$Category_name == "CHECK ME!" &
        aggregated_totals_long$Dollars > 0, na.rm = TRUE)) {
  
  stop("CHECK ME! category contains positive dollars. Investigate the categorization.")
  
} else {
  
  aggregated_totals_long <- aggregated_totals_long |>
    dplyr::filter(!Category_name %in% c("CHECK ME!", "Check Me!"))
}
aggregated_totals_long |> mutate(`Dollars (Millions)` = round(Dollars, digits = 0)) |> select(-Dollars) |>
  select(Year, Category_name, `Dollars (Millions)`, type, Category)

```





```{r}
#| label: tbl-year-totals-table-withgap
#| tbl-cap: "Year totals with Fiscal Futures methodology for excluding/including revenue and expenditure objects."
#| tbl-cap-location: top

year_totals <- aggregated_totals_long |> 
  group_by(type, Year) |> 
  summarize(Dollars = sum(Dollars, na.rm = TRUE)) |> 
  pivot_wider(names_from = "type", values_from = Dollars) |> 

  rename(Expenditures = exp,
         Revenue = rev) |>  
  mutate(`Fiscal Gap` = Revenue - Expenditures)
# creates variable for the Gap each year

year_totals  |>  
  mutate_all(round, digits = 0) |>
  kbl(caption = "Fiscal Gap for each Fiscal Year ($ Millions)") |> 
  kable_styling(bootstrap_options = c("striped"))  |>
  kable_classic() |>   
  add_footnote(c("Values include State CURE dollars (SLFRF)") )

```

Graphs made from `aggregated_totals_long` dataframe.

### Fiscal Gap Graph

```{r}
#| label: fig-fiscal-gap
#| fig-cap: "Fiscal Gap Comparison"
#| fig-subcap: 
#| - "Fiscal Gap With Trend Lines"
#| - "Fiscal Gap Without Trend Lines"
#| fig-cap-location: top

## Adjust x and y coordinates to move placement of text
annotation <- data.frame(
  x = c(2004, 2017, 2019),
  y = c(60, 50, 5),  
  label = c("Expenditures","Revenue", "Fiscal Gap")
)


annotation_nums <- data.frame(
  x = c(2025, 2025, 2025),
  y = c(91, 123, -5),  
  label = c(
   year_totals$Expenditures[year_totals$Year==current_year]/1000,
  year_totals$Revenue[year_totals$Year==current_year]/1000, 
  year_totals$`Fiscal Gap`[year_totals$Year==current_year]/1000)
)

## Dashed line versions for expenditures: 
fiscal_gap <-   
  ggplot(data = year_totals, aes(x=Year, y = Revenue/1000)) +
  geom_recessions(text = FALSE, update = recessions)+

  # geom_smooth adds regression line, graphed first so it appears behind line graph
  geom_smooth(aes(x = Year, y = Revenue/1000), color = "gray", alpha = 0.7, method = "lm", se = FALSE) + 
  #  scale_linetype_manual(values="dashed")+
  geom_smooth(aes(x = Year, y = Expenditures/1000), color ="rosybrown2", linetype = "dotted", method = "lm", se = FALSE, alpha = 0.7) +

  # line graph of revenue and expenditures
  geom_line(aes(x = Year, y = Revenue/1000), color = "Black", size=1) +
  geom_line(aes(x = Year, y = Expenditures/1000, linetype = "dashed"), color = "red", lwd=1) +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000)), color = "darkgray", lwd = 1) +
    geom_hline(yintercept = 0) +

  geom_text(data = annotation, aes(x=x, y=y, label=label,
  parse = TRUE))+

    theme(legend.position = "bottom", legend.title = element_blank())+
    scale_linetype_manual(values = c("dashed", "dashed")) +
  scale_x_continuous(expand = c(0,0),
                                          limits = c(1998, current_year+.5)
) +
  xlab("Year") + 
  ylab("Billions of Dollars")  +
  ggtitle(paste0("Illinois Expenditures and Revenue Totals, 1998-",current_year))

fiscal_gap



## Adjust x and y coordinates to move placement of text
annotation <- data.frame(
  x = c(2024, 2024, 2024),
  y = c(101, 130, 10),  
  label = c("Expenditures","Revenue", "Fiscal Gap")
)


fiscal_gap2 <- ggplot(data = year_totals, aes(x=Year, y = Revenue/1000)) +
  geom_recessions(text = FALSE, update_recessions = recessions)+
  geom_line(aes(x = Year, y = Revenue/1000,  color = "Revenue"), lwd = 1, label = "Revenue") +
  geom_line(aes(x = Year, y = Expenditures/1000, color = "Expenditures"),  linetype = "dotted", lwd = 1, label = "Expenditures") +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000)), color = "darkgray", lwd=1) +
  
  geom_text(data = annotation, aes(x=x, y=y, label=label)) + ## Word locations and text
  geom_text(data = annotation_nums, aes(x = x, y = y, label = scales::dollar(label, accuracy = 0.01L)), size = 3) +  ## Number locations and text
  theme_classic() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  scale_color_manual(values = c("Revenue" = "black", "Expenditures" = "red")) +
    scale_linetype_manual(values = c("Revenue" = "solid", "Expenditures" = "dotted")) + 

  geom_hline(yintercept = 0) +

  scale_y_continuous(#labels = comma, 
                     limits = c(-12, 130), 
                     breaks = c(-10, 20, 40, 60, 80, 100, 120), 
                     minor_breaks = c(-10, 0, 10, 30, 50, 70, 90, 110))+
  scale_x_continuous(expand = c(0,0), 
                     limits = c(1998, current_year+1)
                     ) +
 # scale_color_manual(values = c("red" = "Expenditures", "black" = "Revenue")) + 

  xlab("Year") + 
  ylab("Billions of Dollars")  +
  ggtitle(paste0("Illinois Expenditures and Revenue Totals, 1998-",current_year))


fiscal_gap2

```

```{r}
#| include: false

fiscal_gap2 <- ggplot(data = year_totals, aes(x=Year, y = Revenue/1000)) +
  geom_recessions(text = FALSE, update = recessions)+
  geom_line(aes(x = Year, y = Revenue/1000, color = "Revenue"), lty = "solid"
                ,  lwd = 1) +
  geom_line(aes(x = Year, y = Expenditures/1000, color = "Expenditures"), lty = "dashed"
                , lwd = .7 ) +
  geom_line(aes(x = Year, y = (`Fiscal Gap`/1000), color = "Fiscal Gap"),  lwd=1) +
  
  geom_text(data = annotation, aes(x=x, y=y, label=label))+
  geom_text(data = annotation_nums, aes(x = x+1, y = y, label = scales::dollar(label, accuracy=0.01L)), size = 3) + 
theme_classic() +
theme(
  legend.position = "bottom",
  legend.title = element_blank(),  
  panel.background = element_rect(fill='transparent'), #transparent panel bg
  plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
  panel.grid.major = element_blank(), #remove major gridlines
  panel.grid.minor = element_blank(), #remove minor gridlines
  legend.background = element_rect(fill='transparent'), #transparent legend bg
)+
#scale_linetype_manual(values = c( "dashed", "solid", "solid" )) +
#scale_linetype_manual(labels =c("Expenditures", "Revenue")) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values = c( "Expenditures" = "red", "Revenue" = "black", "Fiscal Gap" = "darkgray"),     
                     guide = guide_legend(override.aes = list(linetype = c( "dashed", "solid", "solid")))
   ) + 
  scale_y_continuous(limits = c(-12, 120), 
                     breaks = c(-10, 20, 40, 60, 80, 100, 120))+
  scale_x_continuous(expand = c(0,0), 
                     limits = c(1998, current_year), 
                     breaks = c(1998, 2000, 2005, 2010, 2015, 2020, current_year) ) +
  xlab("Year") + 
  ylab("Billions of Dollars") 

fiscal_gap2


# ggsave(plot = fiscal_gap2, filename= "./paper-figures/Figure1-fiscalgap.eps", width = 6, height = 4)

#  ggsave(plot = fiscal_gap2, filename= "./paper-figures/Figure1-fiscalgap.png", width = 6, height = 4)

```

## Downloadable Tables


```{r}
#| label: tbl-datahub-export-revenues
#| tbl-cap: "Fiscal Futures Revenue Categories. As of March 2026, this is now the downloadable table on IGPA Datahub site. Totals are in millions of nominal dollars." 


aggregate_rev_labeled <- rev_long |>
  filter(Category_name != "CHECK ME!") |>
  group_by(Year, Category, Category_name) |>
  mutate(Dollars = round(Dollars)) |>
  arrange(Category) |>
  pivot_wider(names_from = "Year", values_from = "Dollars")

datahub_rev <- aggregate_rev_labeled |> select(-type)

datahub_rev |> DT::datatable(rownames = FALSE,
                              extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip', # 'B' for buttons, 'f' for filtering, 'r' for processing, 't' for table, 'i' for info, 'p' for pagination
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), # Specify the download formats
    lengthMenu = list(
      c(10, 25, 50, -1),
      c(10, 25, 50, "All") # Display options for page length
    )
  )
  )
```

```{r}
#| label: tbl-datahub-export-expenditures
#| tbl-cap: "Fiscal Futures Grouped Expenditure Categories. As of March 2026, this is now the downloadable table on IGPA Datahub site. Totals are in millions of nominal dollars. "

aggregate_exp_labeled <- exp_long |>
    filter(Category_name != "CHECK ME!") |>

  group_by(Year, Category, Category_name) |>
  mutate(Dollars = round(Dollars)) |>
  arrange(Category) |>
  pivot_wider(names_from = "Year", values_from = "Dollars")

datahub_exp <- aggregate_exp_labeled |> select(-type)

datahub_exp |> DT::datatable(rownames = FALSE,
                              extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip', # 'B' for buttons, 'f' for filtering, 'r' for processing, 't' for table, 'i' for info, 'p' for pagination
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), # Specify the download formats
    lengthMenu = list(
      c(10, 25, 50, -1),
      c(10, 25, 50, "All") # Display options for page length
    )
  )
  )
```


**Expenditure and revenue amounts in billions of dollars:**

```{r}
#| label: fig-bar-graphs-fy25
#| fig-cap: "FY25 Totals"
#| fig-subcap: 
#| - "FY25 Expenditures"
#| - "FY25 Revenue Sources"

exp_long |>
  filter(Year == current_year & Dollars > 0 ) |>
  arrange(desc(`Dollars`)) |>
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = (`Dollars`/1000), fill = "red"))+ 
  coord_flip() +
      theme_classic()+
  theme(legend.position = "none") +
      labs(title = paste0("Expenditures for ", current_year))+    xlab("Expenditure Categories") +
  ylab("Billions of Dollars") 

rev_long |>
  filter(Year == current_year  & Dollars > 0  ) |>
  arrange(desc(`Dollars`)) |>
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = (`Dollars`/1000)))+ 
  coord_flip() +
    theme_classic() +
    theme(legend.position = "none") +
      labs(title = paste0("Revenue for ", current_year))+
    xlab("Revenue Categories") +
  ylab("Billions of Dollars") 

```

**Expenditure and revenues when focusing on largest categories and combining others into "All Other Expenditures(Revenues)":**

```{r}
#| label: fig-bargraphs-largest-expandrev-cats
#| fig-cap: "Largest Groups"
#| fig-subcap: 
#| - "Largest Expenditures for FY2025"
#| - "Largest Revenue Sources for FY2025"

exp_long |>
  filter( Year == current_year  & Dollars > 0 ) |>
  mutate(rank = rank(Dollars),
        Category_name = ifelse(rank > 13 | Category != "589",  Category_name, 'All Other Expenditures')) |>
 # select(-c(Year, Dollars, rank)) |>
  arrange(desc(Dollars)) |>
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`), y = `Dollars`/1000), fill = "rosybrown2") + 
  coord_flip() +
      theme_classic() +
      labs(title = paste0("Expenditures for ", current_year))+    xlab("") +
  ylab("Billions of Dollars")

rev_long |>
  filter( Year == current_year  & Dollars > 0 ) |>
  mutate(rank = rank(Dollars),
        Category_name = ifelse(rank > 10, Category_name, 'All Other Sources')) |>
  arrange(desc(Dollars)) |>
  ggplot() + 
  geom_col(aes(x = fct_reorder(Category_name, `Dollars`/1000), y = `Dollars`/1000), fill = "dark gray")+ 
  coord_flip() +
      theme_classic() +
      labs(title = paste0("Revenue for ", current_year)) +  
  xlab("") +
  ylab("Billions of Dollars")
```




**Expenditure and revenues when focusing on largest categories and combining others into "All Other Expenditures(Revenues)":**

```{r}
#| label: fig-bargraphs-largest-expend-byfundingsource
#| fig-cap: "Largest Expenditures for FY2025"


n_categories <- 10
# keep top 10 largest category names for bar graph

current_year <- 2025

# list of top 10 expenditure categories
categories <-  exp_long |> 
  filter(Year == 2025 ) |>
  arrange(desc(Dollars)) |>
  slice(1:10)

exp_temp |>
  group_by(year, group, group_name, fund_cat_name) |>
  summarize(sum_expenditure = sum(expenditure)/1000000) |>
  arrange(year) |>
  pivot_wider(names_from = "group_name", values_from = "sum_expenditure") |>
  filter(fund_cat_name == "Federal Trust Funds")


exp_temp |>
  mutate(fund_cat_name =
           case_when(
             fund_cat_name == "General Fund" ~ "General Funds",
             fund_cat_name == "REVOLVING FUNDS" ~ "Revolving Funds",
             T ~ fund_cat_name
           ),
         federal_funded = case_when(
           fund_cat_name == "Federal Trust Funds" ~ "Federal Funds",
           group_name %in% c("MEDICAID", "Medicaid") & fund_cat_name == "General Funds" ~ "Federal Funds",
          T ~ "State Funds"
           
         )) |>
  group_by(year, group, group_name, federal_funded) |>
  summarize(Dollars = sum(expenditure)/1000000) |>
  #filter(fund_cat_name == "Federal Trust Funds")
filter( year == current_year) |>
 # mutate(rank = rank(Dollars),
        mutate(Category_name = ifelse(group %in% categories$Category, group_name, 'All Other Expenditures')) |>
 # select(-c(Year, Dollars, rank)) |>
  arrange(desc(Dollars)) |>
  ggplot() + 
  geom_col(aes(x = Category_name, y = `Dollars`/1000, fill = federal_funded)) + 
  coord_flip() +
      theme_classic() +
      labs(title = paste0("Expenditures for ", current_year))+    xlab("") +
  ylab("Billions of Dollars")

```




### Top 3 Revenues

```{r}
#| label: build-Figure1

annotation <- data.frame(
  x = c(2013, 2018, 2013),
  y = c(16, 10, 5),  
  label = c("Individual Income Tax", "Federal Medicaid", "Sales Tax"
            )
)

top3 <- rev_long  |> 
  filter(Category == "02" | Category == "58" |# Category == "03" | 
           Category == "06") |>
  mutate(Category_name = factor(Category_name, levels = c("Individual Income Taxes", "Federal Medicaid", "Sales Taxes")))

top3 <- ggplot(data = top3, aes(x=Year, y=Dollars/1000))+
  geom_recessions(text = FALSE, update = recessions)+

  geom_line(aes(x=Year, y=Dollars/1000, color = Category_name, lty = Category_name), show.legend = NA, lwd = .7) + 
  #geom_text(data = annotation, aes(x=x, y=y, label=label)) +

  theme_classic() +
  scale_x_continuous(expand = c(0,0),
                     limits = c(1998, current_year+.5)
  ) +
  scale_y_continuous(labels = comma) +
  theme(
  legend.position = c(1, 0.5),
  legend.justification = c(0.5, 0.5),
  base_size = 16,
    plot.margin = margin(5.5, 60, 5.5, 5.5)  # top, right, bottom, left

) +
  # theme(#legend.position = "inside",
  #       #legend.location = "plot",
  #   legend.position = "right",
  #       legend.justification.right = c(1,.3),
  #       #legend.justification.inside = c(1,.3),
  #       base_size = 16) +
  labs( 
       y = "Billions of Nominal Dollars",
       color = NULL,
       lty = NULL) 

Figure1 <- top3  +
  geom_line(data = year_totals, aes(x=Year, y = Revenue/1000, lty = "Total Revenue", color = "Total Revenue"), lwd = 1) +  
  # scale_linetype_manual(values = c("dotted", "dashed", "dotdash", "solid")) +
  scale_color_manual(
  values = c(
    "Individual Income Taxes" = "#1f77b4",
    "Federal Medicaid"        = "#ff7f0e",
    "Sales Taxes"             = "#2ca02c",
    "Total Revenue"           = "black"   
  )) +
 scale_linetype_manual(values = c("dotted", "dashed", "dotdash", "solid"), 
   labels = c( "Individual Income Taxes", "Federal Medicaid", "Sales Taxes", "Total Revenue" )) 


```


```{r}
#| label: fig-FIGURE1
#| fig-cap: "Top 3 Revenue Sources" 



#ggplot2::ggsave("generated/FIGURE1.png", Figure1)#, #width = 11, height = 6, dpi = 300               )
Figure1 


```



```{r}
#| label: fig-top3-ownsourcerevenues
#| fig-cap: "Top 3 Own-Source Revenues Sources"
#| fig-cap-location: top

annotation <- data.frame(
  x = c(2013, 2018, 2013),
  y = c(16, 10, 5),  
  label = c("Individual Income Tax", "Sales Tax", "Corporate Income Tax")
)

top3 <- rev_long  |> 
  filter(Category == "02" | Category == "03" | Category == "06")

top3 <- ggplot(data = top3, aes(x=Year, y=Dollars/1000))+
  geom_recessions(text = FALSE, update = recessions)+

  geom_line(aes(x=Year, y=Dollars/1000, group = Category_name, color = Category_name), position = "stack") + 
 # geom_text(data = annotation, aes(x=x, y=y, label=label)) +

  theme_classic() +
  scale_x_continuous(expand = c(0,0),
                     limits = c(1998, current_year+.5)
  ) +
  scale_y_continuous(labels = comma) +
  scale_linetype_manual(values = c("dotted", "dashed", "solid")) +
  theme(legend.position = "none") +
  labs(title = "Top 3 Own Source Revenues", 
       subtitle = "Individual Income Taxes, Sales Tax, and Corporate income taxes",
       y = "Billions of Nominal Dollars") 


top3  + geom_line(data = year_totals, aes(x=Year, y = Revenue/1000))


```

### Sales Tax

```{r}

adjustments <- readxl::read_xlsx("../Fiscal-Future-Topics/inputs/raw_index_data.xlsx")

salestax <- rev_long  |>
  filter(Category_name == "Sales Taxes") |>
  rename(year = Year, dollars = Dollars)

# salestax <- aggregated_totals_long |> 
#   filter(Category_name == "Sales Taxes")

salestax <- salestax |> left_join(adjustments, by = "year")




# Nominal GROWTH 
sales <- salestax %>% 
  mutate(growth = (dollars/(lag(dollars)) )- 1  )

library(zoo)
library(patchwork)

# 5-year  MOVING AVERAGE 
sales <- sales %>%
  mutate(ma_growth = rollmean(growth, 5, align="center", fill=NA))

#  Calculate averages and plot them  
avg_nom <- mean(sales$growth, na.rm = TRUE)
avg_ma  <- mean(sales$ma_growth,  na.rm = TRUE)


p_points <- ggplot(sales, aes(x = year, y = growth)) +
   geom_recessions(text = FALSE, update = recessions) +

  geom_point(size = 1.6) +
geom_line() + 
  geom_hline(yintercept = avg_nom, linetype = "dashed") +
  scale_y_continuous(labels = percent,# limits = c(-.10, .10),                      
                     breaks = c(-.05, 0, .05, .1)) +
    scale_x_continuous(breaks = c(2000, 2005, 2010, 2015, 2020, 2025))+

  labs(title = "Nominal", x = NULL, y = "Growth rate") +
  theme_minimal(base_size = 12)



p_ma <- ggplot(sales, aes(year, ma_growth)) +
    geom_recessions(text = FALSE, update = recessions)+
    geom_point(size = 1.6) +

  geom_line() +

  geom_hline(yintercept = avg_ma, linetype = "dashed") +
  scale_y_continuous(labels = percent, limits = c(-.05, .10), 
                     breaks = c(-.05, 0, .05, .1)) +
  scale_x_continuous(breaks = c(2000, 2005, 2010, 2015, 2020, 2025))+
  labs(title = "Nominal Moving Average", x = NULL, y = "Moving Average\nGrowth Rate") +
  theme_minimal(base_size = 12)


# Real Dollars Graphs
base_year <- min(salestax$year, na.rm = TRUE)
base_idx  <- salestax %>% filter(year == base_year) %>% 
summarize(x = mean( CPIAUCSL, na.rm = TRUE)) %>% pull(x)

sales <- salestax %>%
mutate(
    real_dollars = dollars * (base_idx / CPIAUCSL),
    r_growth     = real_dollars / lag(real_dollars) - 1,
ma212_real =rollmean(r_growth, 5, align="center", fill=NA)
  )

avg_real   <- mean(sales$r_growth, na.rm = TRUE)
avg_ma_real<- mean(sales$ma212_real, na.rm = TRUE)

p_real <- ggplot(sales, aes(year, r_growth)) +
    geom_recessions(text = FALSE, update = recessions)+

  geom_point(size = 1.6) +
geom_line() + 

 # geom_smooth(se = FALSE, method = "loess", span = 0.6, linewidth = 0.7) +
  geom_hline(yintercept = avg_real, linetype = "dashed") +
  scale_y_continuous(labels = percent,
                     breaks = c(-.05, 0, .05, .1)) +
    scale_x_continuous(breaks = c(2000, 2005, 2010, 2015, 2020, 2025))+

  labs(title = "Real", x = NULL, y = "Real Growth rate") +
  theme_minimal(base_size = 12)

p_ma_real <- ggplot(sales, aes(year, ma212_real)) +
    geom_recessions(text = FALSE, update = recessions)+
    geom_point(size = 1.6) +

  geom_line() +

  geom_hline(yintercept = avg_ma_real, linetype = "dashed") +
  scale_y_continuous(labels = percent, limits = c(-0.05, .1)) +
    scale_x_continuous(breaks = c(2000, 2005, 2010, 2015, 2020, 2025))+

  labs(title = "Real Moving Average", x = NULL, y = "Moving Average\nReal Growth Rate") +
  theme_minimal(base_size = 12)

parts <- c(
  paste("Nominal",         label_percent(accuracy = 0.1)(avg_nom)),
  paste("Real",            if (!is.na(avg_real))    label_percent(accuracy = 0.1)(avg_real)    else NA),
  paste("5-Year MA (Nom.)", label_percent(accuracy = 0.1)(avg_ma)),
  paste("5-Year MA (Real)", if (!is.na(avg_ma_real)) label_percent(accuracy = 0.1)(avg_ma_real) else NA)
)
cap4 <- paste("Dashed lines = sample averages:", paste(na.omit(parts), collapse = "; "))


p_sales4 <- (p_points | p_real) / (p_ma | p_ma_real) +
  plot_annotation(
    title = "Various Measures of Illinois Sales Tax Growth Rate",
    caption = cap4
  )

print(p_sales4)

#ggsave("generated/figure2.png", p_sales4#, width = 12, height = 8, dpi = 300)


```

### Own Source and Fed Transfers

Federal revenue to all fund types is coded as 3 types of federal revenue: Medicaid, Transportation, and Other. 

```{r}
#| label: tbl-ownsource-and-fed-revenue
#| tbl-cap: "Own Source and Federal Revenue"
#| tbl-location: margin
 

ownsource_rev <- rev_long |>
  filter(!Category %in% c("57", "58", "59")) |>
  group_by(Year) |> 
  summarize(Dollars = sum(Dollars))

ownsource_rev
# ownsource_rev |> 
#   ggplot()+geom_line(aes(x=Year, y=Dollars)) + 
#   labs(title = "Own Source Revenues", subtitle = "Total own source revenue", y = "Millions of Dollars")

```

```{r}
fed_rev <- ff_rev |> 
  select(fy, rev_57, rev_58, rev_59) |>
  mutate(fed_total = rev_57+rev_58+rev_59)


annotation <- data.frame(
  x = c(2013, 2015),
  y = c(50, 25),  
  label = c("Own Source Revenue", "Federal Revenue")
)

annotation_nums <- data.frame(
  x = c(2023, 2023),
  y = c(50, 25),  
  label = c("Own Source Revenue", "Federal Revenue")
)

ownsource_graph <- ggplot(ownsource_rev, aes(x=Year, y=Dollars/1000)) + 
  geom_recessions(text = FALSE, update_recessions = recessions)+
  geom_line(data = ownsource_rev, aes(x=Year, y=Dollars/1000), color = "Red") + 
  geom_line(data = fed_rev, aes(x=fy, y=fed_total/1000), color = "Black") + 
    geom_text(data = annotation, aes(x=x, y=y, label=label))+
    scale_y_continuous(labels = comma)+
  scale_x_continuous(expand = c(0,0),
                     limits = c(1998-.5, current_year+.5)) + 
  theme(legend.position = "none") +

  theme_classic()+
  labs(title = "Own Source Revenue and Federal Revenue, 1998-2025", 
  y = "Billions of Dollars")
ownsource_graph

```


### Medicaid Reimbursements and Expenditures

Key funds: Healthcare Provider Relief (0793) and Hospital Provider (0346)
0365 is Health and Human Services Medicaid TR ? 
0740 is Medicaid Buy in Program

Healthcare provider taxes come from revenue source 0133.

2104 = Medicare Part D
2683 = MCO Provider Assessment --> Mostly goes to Healthcare Provider Relief Fund 0793 in HFS
2526 = Hospital Provider Fund (not used in fy2025)

0133 = Health Care Provider Tax --> Mostly goes to Hospital Provider Fund 0346
0137 = Health Care Prov-Hospital (not used in  fy2025)
0145 = IHFA Medicaid Provider (not used in fy2025)


**Illinois also received:**   

- over 500 million for TANF grant (source = 1393).

- Also receives food stamp funds, over 500 million for "Medical Administration" source 0675. This mostly goes to the General Revenue Fund.  

- around \$1.3 billion for "Medical Assistance" (source 0692, which goes to Drug Rebate Fund 0728)

- an additional $3.3 billion in medical assistance from source 0676 which mostly goes to the General Revenue fund (0001) and a little goes to Tobacco settlement recovery fund


```{r}
provider_taxes <- rev_temp |> 
  filter(source == "0133" | source == "2683" ) |>
  group_by(fy, source) |> 
  summarize(receipts=sum(receipts)) 

ggplot()+
  geom_line(data = provider_taxes, aes(x=fy, y = receipts, color = source)) +
 labs(title = "Provider Taxes", caption = "Revenue Sources 0133 and 2683.") +
  scale_y_billions(name = "Dollars")

rev_temp |>
  filter(fy == 2025) |>
  filter(source == "0133") |>
  group_by(fy, fund_name_ab) |> 
  summarize(receipts=sum(receipts))

exp_temp |> filter(fund == "0346" & fy == 2025) |> summarize(spent = sum(expenditure))
exp_temp |> filter(fund == "0346" & fy == 2025)

rev_temp |>
  # Healthcare Provider Taxes & MCO Provider assessment
  filter(source == "0133" | source == "2683") |>
  group_by(fy, fund_name_ab) |> 
  summarize(receipts=sum(receipts)) |> 
  ggplot() + 
  geom_col(aes(x=fy, y = receipts, fill = fund_name_ab)) + 
  scale_y_billions() + 
  labs(x = element_blank(),
       title = "Funds that received Provider Taxes")



medicaid_cost <- exp_temp |> 
  filter(agency=="478" & 
           #(appr_org=="01" | appr_org == "65" | appr_org=="88") &
           (object=="4900" | object=="4400")) |> group_by(fy) |> 
  summarize(sum=sum(expenditure, na.rm=TRUE))


med_reimburse <- rev_temp |> 
  filter(rev_type == "58") |>
   # rev_type=="57" & agency=="478" & 
         #  (source=="0618"|source=="2364"|
           #   source=="0660"|source=="1552"| source=="2306"| source=="2076"|source=="0676"|source=="0692") ) |> 
  group_by(fy) |> summarize(sum=sum(receipts, na.rm=TRUE))


ggplot()+
  geom_line(data=medicaid_cost, aes(x=fy, y=sum), color = "red") + 
  geom_line(data=med_reimburse, aes(x=fy, y = sum), color="black") +
 # geom_line(data = provider_taxes, aes(x=fy, y = receipts), color = "gray") +
 labs(title = "Medicaid reimbursements and Medicaid expenditures", 
      caption = "Medicaid expenditures include funds provided to medical providers.",
      x = element_blank()) +
  scale_y_billions(name = "Dollars") + 
  scale_x_continuous(breaks = c(2000, 2005, 2010, 2015, 2020, 2025))
```


### Provider Assessments, % of 1998 levels

```{r}
#| label: fig-FIGURE3
##| fig-width: 6
##| fig-height: 4


provider_taxes <- rev_long |> filter(Category == "31")

base_idx  <- provider_taxes %>% filter(Year == base_year)


provider_taxes <- provider_taxes |>
mutate(rev_indexed = (Dollars / Dollars[Year==1998]),
Series = "Provider Taxes") |>
select(Year, rev_indexed, Series)

rev_index = year_totals |>
mutate(rev_indexed = (Revenue / Revenue[Year==1998]),
Series = "Total Revenue") |>
select(Year, rev_indexed, Series)

plotdata <- rbind(provider_taxes, rev_index)

plot_indexed <- 
ggplot(plotdata, aes(Year, rev_indexed, color = Series)) + 
     geom_recessions(text = FALSE, update = recessions) +

geom_line(lwd = 1) +
 geom_point(data = plotdata |> group_by(Series) |> filter(Year == max(Year)),
    aes(color = Series),
    hjust = -0.1, vjust = 0.5, show.legend = FALSE ) + 
    scale_color_manual(values = c("#2A5783",  "#99C5E3") )+

  geom_label(data = plotdata |> group_by(Series) |> filter(Year == max(Year)),
    aes(label = scales::percent(rev_indexed , accuracy = 1), color = Series),
    hjust = -0.1, vjust = 0.5, show.legend = FALSE ) +
scale_y_continuous(labels = scales::percent, limits = c(0,9), name = NULL) + 
scale_x_continuous(limits = c(base_year, 2025 + 2), breaks = c(2000, 2005, 2010, 2015, 2020,2025), name = NULL) + 

  #labs(
   #title = "Medical Provider Assessments, as % of 1998 levels") +
theme(base_size = 16,
legend.position = "bottom")

#ggsave(plot_indexed, file= "generated/figure3.png")

#ggsave(plot_indexed, file= "generated/figure3_notitle.png")

plot_indexed

```

```{r}
#| label: fig-medicaid-obbb-plot
##| fig-width: 6
##| fig-height: 4
#| fig-cap: "Projected State Revenue Losses from OBBB Caps on Provider Tax Rates (MCO & Hospital Assessments Only). Source: University of Illinois Fiscal Futures Project."
#| fig-cap-location: margin

#obbb_raw <- readxl::read_excel("../Fiscal-Future-Topics/data/FY2025 Files/Medicaid/obbb_v6.xlsx")
obbb_raw <- readxl::read_excel("../Fiscal-Future-Topics/data/FY2025 Files/Medicaid/obbb_v9.xlsx")


obbb_raw <- obbb_raw |> 
  mutate(Scenario = case_when(
    Scenario == "Baseline:  no growth in patient revenues" ~ "Baseline: no growth",
       Scenario == "Modest growth in patient revenues" ~ "Modest Growth", 
       Scenario == "Modest contraction in patient revenues" ~ "Modest Contraction"
))

obbb_tidy <- obbb_raw |>
pivot_longer(`2025`: `2033`, names_to = "Year", values_to = "value") |> 
  mutate(Year = as.numeric(Year))

p_obbb <- ggplot(obbb_tidy, aes(x = Year, y = value, group=Scenario, color = Scenario)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_label(data = obbb_tidy |> group_by(Scenario) |> filter(Year == max(Year)),
            aes(label = scales::dollar(value , accuracy = 1)),
            hjust = -0.1, vjust = 0.5, show.legend = FALSE )+
  
  scale_color_manual(values = c("#2A5783", "#6495BF", "#B9DDF1") )+
  scale_y_continuous(labels = dollar, name ="Millions") +
  scale_x_continuous(limits = c(2026, 2034), breaks = c(2026:2033), name = NULL) + 
  
  theme(legend.position = "bottom",
        base_size = 16) #+ 
# labs(title= "Projected State Revenue Losses\nfrom OBBB's Caps on Provider Tax Rates",
#subtitle = "Based on MCO & Hospital Assessments Only")

#ggsave(p_obbb, file= "generated/figure4.png")
#ggsave(p_obbb, file= "generated/figure4_notitle.png")

p_obbb

```


# Change from Previous Year

```{r}
#| fig-height: 6

rev_long |>
  filter(Year == current_year | Year == past_year) |>
  mutate(Year = as.character(Year)) |>
  ggplot(aes(x = Dollars, y = reorder(Category, Dollars))) +
  geom_line(aes(group = Category) )+
  geom_text(aes(x = ifelse(Year == current_year, as.numeric(Dollars), NA),  label = ifelse(Year == current_year, Category_name, "")),  
            hjust = -0.2,
            size = 2.8) +
  geom_point(aes(color = Year), size=2)  +
  labs(title = "2024 to 2025 Change in Revenue", x = "Millions of Dollars" , y = "",  caption = "")  +
  scale_fill_manual(values = c("#d62828", "#003049"), labels = c("FY 2024", "FY 2025"))+
  scale_color_manual(values = c("#d62828", "#003049")) +   
  theme_classic()+ 
  theme(
    legend.position = "bottom" ,
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y.left  = element_blank()) +
  scale_x_continuous(
    limits = c(0, 40000), 
                     labels = comma)

exp_long |>
  filter(!Category %in%  c("Check Me!", "CHECK ME") ) |>
  filter(Year == current_year | Year == past_year) |>
  mutate(Year = as.character(Year)) |>
  ggplot(aes(x = Dollars, y = reorder(Category, Dollars))) +
  geom_line(aes(group = Category) )+
  geom_text(aes(x = ifelse(Year == current_year, as.numeric(Dollars), NA),  
                label = ifelse(Year == current_year, Category_name, "")),  
            hjust = -0.2,
            size = 2.8) +
  geom_point(aes(color = Year), size=2)  +
  labs(title = "2024 to 2025 Change in Expenditures", x = "Millions of Dollars" , y = "",  caption = "")  +
  scale_fill_manual(values = c("#d62828", "#003049"), labels = c("FY 2024", "FY 2025"))+
  scale_color_manual(values = c("#d62828", "#003049")) +
  theme_classic()+ 
  theme(
    legend.position = "bottom" ,
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y.left  = element_blank())+
  scale_x_continuous(limits = c(0, 40000), 
                     labels = comma
                     )
```

Each year, you will need to update the CAGR formulas! Change the filter() year.

`calc_cagr` is a function created for calculating the CAGRs for different spans of time.

```{r}
exp_totals <- ff_exp |> 
  rowwise() |> 
  mutate(exp_TOTALS = sum(across(exp_402:exp_970), na.rm=TRUE))

rev_totals <- ff_rev |>    
  rowwise() |> 
  mutate(rev_TOTALS = sum(across(rev_02:rev_78), na.rm=TRUE))



rev_long <- pivot_longer(rev_totals, rev_02:rev_TOTALS, names_to = c("type","Category"), values_to = "Dollars", names_sep = "_") |> 
  rename(Year = fy) |>
  mutate(Category_name = case_when(
    Category == "02" ~ "INDIVIDUAL INCOME TAXES" ,
    Category == "03" ~ "CORPORATE INCOME TAXES" ,
    Category == "06" ~ "SALES TAXES" ,
    Category == "09" ~ "MOTOR FUEL TAX" ,
    Category == "12" ~ "PUBLIC UTILITY TAXES" ,
    Category == "15" ~ "CIGARETTE TAXES" ,
    Category == "18" ~ "LIQUOR GALLONAGE TAXES" ,
    Category == "21" ~ "INHERITANCE TAX" ,
    Category == "24" ~ "INSURANCE TAXES, FEES & LICENSES" ,
    Category == "27" ~ "CORP FRANCHISE TAXES & FEES" ,
    Category == "30" ~ "HORSE RACING TAXES & FEES",  # in Other
    Category == "31" ~ "MEDICAL PROVIDER ASSESSMENTS" ,
    Category == "32" ~ "GARNISHMENT-LEVIES" , # dropped
    Category == "33" ~  "LOTTERY RECEIPTS" ,
    Category == "35" ~  "OTHER TAXES" ,
    Category == "36" ~  "RECEIPTS FROM REVENUE PRODUCING", 
    Category == "39" ~  "LICENSES, FEES & REGISTRATIONS" ,
    Category == "42" ~  "MOTOR VEHICLE AND OPERATORS" ,
    Category == "45" ~  "STUDENT FEES-UNIVERSITIES",   # dropped
    Category == "48" ~  "RIVERBOAT WAGERING TAXES" ,
    Category == "51" ~  "RETIREMENT CONTRIBUTIONS" , # dropped
    Category == "54" ~ "GIFTS AND BEQUESTS", 
    Category == "57" ~  "FEDERAL OTHER" ,
    Category == "58" ~  "FEDERAL MEDICAID", 
    Category == "59" ~  "FEDERAL TRANSPORTATION" ,
    Category == "60" ~  "OTHER GRANTS AND CONTRACTS", #other
    Category == "63" ~  "INVESTMENT INCOME", # other
    Category == "66" ~ "PROCEEDS,INVESTMENT MATURITIES" , #dropped
    Category == "72" ~ "BOND ISSUE PROCEEDS",  #dropped
    Category == "75" ~  "INTER-AGENCY RECEIPTS ",  #dropped
    Category == "76" ~  "TRANSFER IN FROM OUT FUNDS",  #other
    Category == "78" ~  "ALL OTHER SOURCES" ,
    Category == "79" ~   "COOK COUNTY IGT", #dropped
    Category == "98" ~  "PRIOR YEAR REFUNDS", #dropped
    Category == "TOTALS" ~ "Total",
    T ~ "CHECK ME"
    
  ) ) |> 
  select(-type, -Category) |>  # drop extra columns type and Category number
  group_by(Year, Category_name) |>
  summarise(Dollars= round(sum(Dollars, na.rm=TRUE), digits=2)) |> 
  mutate(Category_name = str_to_title(Category_name))


# creates wide version of table where each revenue source is a column
revenue_wide2 <- rev_long |> 
  pivot_wider(names_from = Category_name, 
              values_from = Dollars) |>
  #  relocate("Other Revenue Sources **", .after = last_col()) |>
  relocate("Total", .after =  last_col())

```


```{r}

exp_long <- pivot_longer(exp_totals, exp_402:exp_TOTALS , 
                         names_to = c("type", "Category"), values_to = "Dollars", names_sep = "_") |>
  rename(Year = fy ) |> 
  mutate(Category_name = 
           case_when(
             Category == "402" ~ "Aging" ,
             Category == "406" ~ "Agriculture", 
             Category == "416" ~ "Central Management",
             Category == "418" ~ "Children & Family Services", 
             Category == "420" ~ "Commerce and Economic Development",
             Category == "422" ~ "Natural Resources" ,
             Category == "426" ~ "Corrections",
             Category == "427" ~ "Employment Security" ,
           #  Category == "442" ~ "Human Rights" ,
             Category == "442" ~ "Other Departments" ,

             Category == "444" ~ "Human Services" ,
            
           # Category == "445" ~ "IL Power Agency" ,            
            Category == "445" ~ "Other Departments" ,
           
           #  Category == "452" ~ "Labor" ,
              Category == "452" ~ "Other Departments" ,

             Category == "458" ~ "State Lottery" ,
             
             Category == "478" ~ "Healthcare & Family Services (Net Medicaid)", 
             Category == "482" ~ "Public Health", 
             Category == "492" ~ "Revenue", 
             Category == "494" ~ "Transportation" ,
                       
            Category == "497" ~ "Veterans' Affairs" ,

             Category == "497" ~ "Other Departments" ,
             Category == "507" ~ "Other Departments",  
            # Category == "507" ~ "GOMB",

             Category == "532" ~ "Environmental Protection Agency" ,
           
             Category == "557" ~ "Tollway" ,
             Category == "589" ~ "State Employment Retirement (SERS)",

             
           Category == "901" ~ "State Pension Contribution",
            
            Category == "903" ~ "Debt Service",
             Category == "904" ~ "State Employee Healthcare",
             Category == "910" ~ "Legislative"  ,
             Category == "920" ~ "Judicial" ,
             Category == "930" ~ "Elected Officials" , 
             Category == "941" ~ "Public Safety" ,
             Category == "942" ~ "Econ Dev & Infrastructure" ,
             Category == "943" ~ "Central Services",
             Category == "944" ~ "Business & Professional Regulation" ,
             Category == "945" ~ "Medicaid" ,
             Category == "946" ~ "Capital Improvement" , 
             Category == "948" ~ "Other Departments" ,
             Category == "949" ~ "Other Boards & Commissions" ,
             Category == "959" ~ "K-12 Education" ,
             Category == "960" ~ "University Education",
             Category == "970" ~ "Local Govt Revenue Sharing",
             Category == "TOTALS" ~ "Total",
             T ~ "CHECK ME" 
            #  T ~ "All Other Expenditures **")
           )) |> 
    select(-type, -Category) |> 

  group_by(Year, Category_name) |> 
  summarise(Dollars = round(sum(Dollars, na.rm=TRUE), digits=2))

expenditure_wide2 <- exp_long |> 
  pivot_wider(names_from = Category_name, 
              values_from = Dollars) |>
  relocate("Total", .after =  last_col())



```

Things to do when updating the code:

-   Each year, you need to increase the cagr value by 1. The value should be the (current year - 1998). For FY23, this is 2023-1998 = 25. So all cagr values that were 24 will be changed to 25.


```{r}

max_cagr_years = current_year-1998

# function for calculating the CAGR
calc_cagr <- function(df, n) {
  df <- df |>
    arrange(Category_name, Year) |>
    group_by(Category_name) |>
    mutate(cagr = ((`Dollars` / lag(`Dollars`, n)) ^ (1 / n)) - 1,
           cagr = ifelse(is.na(cagr), 0, cagr)
)

  return(df)
}

cagr_calculations <- function(df){ # This works for one variable at a time
  
  df <- df
  
  cagr_max <- calc_cagr(df, max_cagr_years) |> 
    summarize(cagr_max = round(sum(cagr*100, na.rm = TRUE), 2))

  # Update year in the filter() and summarize() commands to current year.
  cagr_10 <- calc_cagr(df, 10) |> 
    filter(Year == current_year) |>
    summarize(cagr_10 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_5 <- calc_cagr(df, 5) |> 
    filter(Year == current_year) |>
    summarize(cagr_5 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_3 <- calc_cagr(df, 3) |> 
    filter(Year == current_year) |>
    summarize(cagr_3 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_2 <- calc_cagr(df, 2) |> 
    filter(Year == current_year) |>
    summarize(cagr_2 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))

  cagr_1 <- calc_cagr(df, 1) |> 
    filter(Year == current_year) |>
    summarize(cagr_1 = case_when(Year == current_year ~ round(sum(cagr*100, na.rm = TRUE), 2)))
  
    # Combine all into one tibble
  result <- data.frame(cagr_max, cagr_10, cagr_5, cagr_3, cagr_2, cagr_1)
  return(result)
}

```


```{r}
#| label: tbl-good-exp-CAGR-withTotals
#| tbl-cap: "Expenditure Category CAGRs with Total CAGR (Ordered Alphabetically)"
#| tbl-cap-location: top


CAGR_expenditures_summary_tot <- cagr_calculations(exp_long)   |>
  filter(Category_name != "CHECK ME") |>     #  You should havec checked that there were no unlabeled expenditures or revenues way before this point!! 
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) |> 
  rename("Expenditure Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10, "27 Year CAGR" = cagr_max )


totalrow <- which(grepl("Total", CAGR_expenditures_summary_tot$`Expenditure Category`))

CAGR_expenditures_summary_tot <- move_to_last(CAGR_expenditures_summary_tot, totalrow) 

lastrow = nrow(CAGR_expenditures_summary_tot)


CAGR_expenditures_summary_tot |>   
  kbl(caption = "CAGR Calculations for All Expenditure Categories" , row.names=FALSE) |> 
  kable_classic() |>
  row_spec(lastrow, bold = T, color = "black", background = "gray")

```

```{r}
#| label: tbl-RevCAGRs-allcats
#| tbl-cap: "Revenue Category CAGRs with Total CAGR (Ordered Alphabetically)"
#| tbl-cap-location: top

CAGR_revenue_summary_tot <- cagr_calculations(rev_long) |>     
  filter(Category_name != "CHECK ME") |>     #  You should havec checked that there were no unlabeled expenditures or revenues way before this point!!  
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) |> 
  rename("Revenue Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10, "27 Year CAGR" = cagr_max )

CAGR_revenue_summary_tot <- move_to_last(CAGR_revenue_summary_tot, 1)

totalrow <- which(grepl("Total", CAGR_revenue_summary_tot$`Revenue Category`))

CAGR_revenue_summary_tot <- move_to_last(CAGR_revenue_summary_tot, totalrow)

lastrow = nrow(CAGR_revenue_summary_tot)

CAGR_revenue_summary_tot |> 
  kbl(caption = "CAGR Calculations for All Revenue Sources (Ordered Alphabetical)", row.names = FALSE) |> 
   kable_classic() |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")
```


```{r}
#| label: tbl-AppendixItem1
#| tbl-cap: "Yearly Change in Revenues - All FF Categories, Ordered from Largest to Smallest Revenue Amount"


first_year = as.numeric(1998)
n_year_change = as.numeric(current_year-1998)



revenue_change2 <- rev_long |>
  filter(Year >= past_year | Year == first_year) |>
    filter(Category_name != "CHECK ME") |>     #  You should havec checked that there were no unlabeled expenditures or revenues way before this point!! 
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") |>
  rename(    
    Dollars_current = Dollars_2025,
    Dollars_lastyear = Dollars_2024
    )|>
  mutate(
    "Current FY ($ billions)" = round(Dollars_current/1000, digits = 2),
    "Past FY ($ billions)" = round(Dollars_lastyear/1000, digits = 2),
        "FY 1994 ($ billions)" = round(Dollars_1998/1000, digits = 2),
    "1-Year Change" = percent((Dollars_current -Dollars_lastyear)/Dollars_lastyear, accuracy = .01)) |>
  left_join(CAGR_revenue_summary_tot, by = c("Category_name" = "Revenue Category")) |> 
    arrange(-`Current FY ($ billions)`)|>
    mutate(`27 Year CAGR` = percent(`27 Year CAGR`/100, accuracy=.01)) |>

  rename(  
          "Revenue Category" = Category_name ) |>
  select(-c( Dollars_1998, Dollars_current, Dollars_lastyear, `1 Year CAGR`:`10 Year CAGR`))


allother_row <- which(grepl("All Other", revenue_change2$`Revenue Category`))

revenue_change2 <- move_to_last(revenue_change2, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", revenue_change2$`Revenue Category`))
revenue_change2 <- move_to_last(revenue_change2, totalrow) # Move "Total" to last row

lastrow = nrow(revenue_change2)

Table2 <- revenue_change2 |>   
  filter(!is.na(`Revenue Category`)) |>

  kbl(caption = "Table 1. Yearly Change in Revenue (All Sources)", row.names = FALSE) |> 
   kable_classic() |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")



#save_kable(Table2, file = "generated/App1_AllRevenueSources.html", self_contained = TRUE)

Table2
```


```{r}
#| label: tbl-AppendixItem2
#| tbl-cap: "Yearly Change in Expenditures - All FF Categories, Ordered from Largest to Smallest Expenditure Amount"

expenditure_change2 <- exp_long |>
    filter(Category_name != "CHECK ME") |>     #  You should havec checked that there were no unlabeled expenditures or revenues way before this point!! 
  group_by(Year, Category_name) |>
  summarize(Dollars = sum(Dollars, na.rm=TRUE)) |>
  ungroup() |>
  filter(Year >= past_year | Year == first_year ) |>
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") |>
    rename(    
    Dollars_current = Dollars_2025,
    Dollars_lastyear = Dollars_2024
    )|>
  mutate(
    "FY 2025 ($ billions)" = round(Dollars_current/1000, digits = 2),
    "FY 2024 ($ billions)" = round(Dollars_lastyear/1000, digits = 2),
    "FY 1998 ($ billions)" = round(Dollars_1998/1000, digits = 2),
    "1-Year Change" = percent((Dollars_current -Dollars_lastyear)/Dollars_lastyear, accuracy = .01)) |>
  left_join(CAGR_expenditures_summary_tot, by = c("Category_name" = "Expenditure Category")) |> 
  arrange(-`FY 2025 ($ billions)`)|>
    mutate(`27 Year CAGR` = percent(`27 Year CAGR`/100, accuracy=.01)) |>

  select(-c( Dollars_1998, Dollars_current, Dollars_lastyear, `1 Year CAGR`:`10 Year CAGR`)) |>

  rename("Expenditure Category" = Category_name ) # |> filter(!is.na(`Expenditure Category`))


allother_row <- which(grepl("All Other", expenditure_change2$`Expenditure Category`))

expenditure_change2 <- move_to_last(expenditure_change2, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", expenditure_change2$`Expenditure Category`))
expenditure_change2 <- move_to_last(expenditure_change2, totalrow) # Move "Total" to last row

lastrow = nrow(expenditure_change2)

Table3<- expenditure_change2 |> 
  kbl(row.names = FALSE) |> 
  kable_classic() |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")

  

#save_kable(Table3, file = "generated/App2_AllExpenditureCategories.html", self_contained = TRUE)
Table3         
```

## Summary Tables - Largest Categories

The 10 largest revenue sources and 15 largest expenditure sources remain separate categories and all other smaller sources/expenditures are combined into "All Other Revenues (Expenditures)". These condensed tables are typically used in the Fiscal Futures articles. They were manually created in past years but this hopefully automates the process a bit until final formatting stages.


```{r}
#| label: tbl-top-rev-CAGRs
#| tbl-cap: Largest Revenue Categories with CAGRs

n_categories <- 10+1 # (Top 10 and then Total )

categories <-  rev_long |> 
  filter(Year == current_year  ) |>
  arrange(desc(Dollars)) |>
  slice(1:n_categories)

rev_majorcats <- rev_long |> 
  filter(
    (Year == current_year | Year == first_year)
          & Category_name %in% categories$Category_name) 

rev_long_majorcats <- rev_long |>
  mutate(Category_name = ifelse(Category_name %in% rev_majorcats$Category_name, Category_name, "All Other Sources"),
         Category_name = ifelse(Category_name == "Total", "Total Revenue", Category_name)) |>
  group_by(Year, Category_name) |> 
  summarize(Dollars = sum(Dollars, na.rm=TRUE))


# creates wide version of table where each revenue source is a column
revenue_wide_majorcats <- rev_long_majorcats |> 
  pivot_wider(names_from = Category_name, 
              values_from = Dollars) |>
  relocate("All Other Sources", .after = last_col()) |>
  relocate("Total Revenue", .after =  last_col())
```

```{r}
#| label: tbl-top10-rev-CAGRs
#| tbl-cap: Top 10 Revenue Sources with CAGRs
#| tbl-cap-location: top


CAGR_revenue_majorcats_tot <- cagr_calculations(rev_long_majorcats) |>   
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) |> 
  rename("Revenue Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10, "27 Year CAGR" = cagr_max )

allother_row <- which(grepl("All Other", CAGR_revenue_majorcats_tot$`Revenue Category`))

CAGR_revenue_majorcats_tot <- move_to_last(CAGR_revenue_majorcats_tot, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", CAGR_revenue_majorcats_tot$`Revenue Category`))
CAGR_revenue_majorcats_tot <- move_to_last(CAGR_revenue_majorcats_tot, totalrow) # Move "Total" to last row

lastrow = nrow(CAGR_revenue_majorcats_tot)

CAGR_revenue_majorcats_tot |> 
  kbl(caption = "CAGR Calculations for Largest Revenue Sources", row.names = FALSE) |> 
  kable_classic() |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")
```

```{r}
#| label: build-Table1

### Yearly change summary table for Top 10 Revenues ###
revenue_change_majorcats <- rev_long_majorcats |>
  filter(Year >= past_year |  Year == first_year) |>
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") |>
    rename(    
    Dollars_current = Dollars_2025,
    Dollars_lastyear = Dollars_2024
    )|>
  mutate(
    "Current FY ($ billions)" = round(Dollars_current/1000, digits = 2),
    "Previous FY ($ billions)" = round(Dollars_lastyear/1000, digits = 2),
    "FY 1998 ($ billions)" = round(Dollars_1998/1000, digits = 2),
    "1-Year Change" = percent((Dollars_current -Dollars_lastyear)/Dollars_lastyear, accuracy = .01),
  )  |>
  left_join(CAGR_revenue_majorcats_tot, by = c("Category_name" = "Revenue Category")
            ) |> 
  arrange(-`Current FY ($ billions)`)|>
  mutate(`27 Year CAGR` = percent(`27 Year CAGR`/100, accuracy=.01)) |>
  select(-c(Dollars_1998, Dollars_current, Dollars_lastyear, `1 Year CAGR`:`10 Year CAGR`
            )) |>
  rename("Revenue Category" = Category_name )

allother_row <- which(grepl("All Other", revenue_change_majorcats$`Revenue Category`))

revenue_change_majorcats <- move_to_last(revenue_change_majorcats, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", revenue_change_majorcats$`Revenue Category`))
revenue_change_majorcats <- move_to_last(revenue_change_majorcats, totalrow) # Move "Total" to last row

lastrow = nrow(revenue_change_majorcats)

Table1 <- revenue_change_majorcats|> 
  kbl(caption = "Yearly Change in Revenue for Main Revenue Sources", row.names = FALSE, align = "l") |> 
   kable_classic() |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")

```



```{r}
#| label: tbl-TABLE1
#| tbl-cap: "Top 10 Revenue Sources with CAGRs"


#save_kable(Table1, file = "generated/TABLE1.html", self_contained = TRUE)

Table1


```



```{r}
#| label: tbl-top-exp-CAGRs
#| tbl-cap: Largest Expenditure Categories with CAGRs

n_categories <- 10 + 1 # (Top 9 and then Total )

# keep top 10 largest categories or categories larger than 2 billion for final table in report (not a set rule, changes each year depending what the focus of the report is or what is highlighted.)

categories <-  exp_long |> 
  filter(Year == current_year & Category_name != "State Employment Retirement (SERS)" ) |>
  arrange(desc(Dollars)) |>
  slice(1:n_categories)

exp_majorcats <- exp_long |> 
  filter(
    (Year == current_year | Year == first_year)
          & Category_name %in% categories$Category_name) 


exp_long_majorcats <- exp_long |>
  mutate(Category_name = ifelse(Category_name %in% exp_majorcats$Category_name, Category_name, "All Other Expenditures **"),
         Category_name = ifelse(Category_name == "Total", "Total Expenditures", Category_name)) |>
  group_by(Year, Category_name) |> 
  summarize(Dollars = sum(Dollars, na.rm=TRUE))

# expenditure_wide_majorcats <- exp_long_majorcats |> 
#   pivot_wider(names_from = Category_name, 
#               values_from = Dollars) |>
#   relocate("All Other Expenditures **", .after = last_col()) |>
#   relocate("Total Expenditures", .after =  last_col())



# CAGR values for largest expenditure categories and combined All Other Expenditures


CAGR_expenditures_majorcats_tot <- cagr_calculations(exp_long_majorcats) |>
  select(-c(Category_name.1, Category_name.2, Category_name.3, Category_name.4, Category_name.5 )) |> 
  rename("Expenditure Category" = Category_name, "1 Year CAGR" = cagr_1, "2 Year CAGR" = cagr_2, "3 Year CAGR" = cagr_3, "5 Year CAGR" = cagr_5, "10 Year CAGR" = cagr_10,"27 Year CAGR" = cagr_max )

allother_row <- which(grepl("Other", CAGR_expenditures_majorcats_tot$`Expenditure Category`))

CAGR_expenditures_majorcats_tot <- move_to_last(CAGR_expenditures_majorcats_tot, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", CAGR_expenditures_majorcats_tot$`Expenditure Category`))
CAGR_expenditures_majorcats_tot <- move_to_last(CAGR_expenditures_majorcats_tot, totalrow) # Move "Total" to last row

lastrow = nrow(CAGR_expenditures_majorcats_tot)

CAGR_expenditures_majorcats_tot|>   
  kbl(caption = "CAGR Calculations for Largest Expenditure Categories" , row.names=FALSE) |> 
     kable_classic()  |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")


# Yearly change for Top n largest expenditure categories
expenditure_change_majorcats <- exp_long_majorcats |>
  filter(Year >= past_year |
           Year == first_year & Category_name != "SERS") |>
  pivot_wider(names_from = Year , values_from = Dollars,   names_prefix = "Dollars_") |>   
  rename(    
    Dollars_current = Dollars_2025,
    Dollars_lastyear = Dollars_2024
    )|>
  mutate("Current FY ($ Billions)" = round(Dollars_current/1000, digits = 2),
         "Previous FY ($ Billions)" = round(Dollars_lastyear/1000, digits = 2),
         "FY 1998 ($ Billions)" = round(Dollars_1998/1000, digits = 2),
         "1-Year Change" = percent((Dollars_current -Dollars_lastyear)/Dollars_lastyear, accuracy = .01),
         )  |>
 left_join(CAGR_expenditures_majorcats_tot, by = c("Category_name" = "Expenditure Category")) |> 
  arrange(-`Current FY ($ Billions)`)|>
  mutate(`27 Year CAGR` = percent(`27 Year CAGR`/100, accuracy=.01)) |>
  select(-c(Dollars_1998, Dollars_current, Dollars_lastyear, `1 Year CAGR`:`10 Year CAGR`
            )) |>
  rename(
   # "1-Year Change" = `1 Year CAGR`,
    "27 Year Change" = `27 Year CAGR`, 
          "Expenditure Category" = Category_name )

allother_row <- which(grepl("All Other", expenditure_change_majorcats$`Expenditure Category`))

expenditure_change_majorcats <- move_to_last(expenditure_change_majorcats, allother_row) # Move "All Other" to 2nd to last row

totalrow <- which(grepl("Total", expenditure_change_majorcats$`Expenditure Category`))
expenditure_change_majorcats <- move_to_last(expenditure_change_majorcats, totalrow) # Move "Total" to last row

lastrow = nrow(expenditure_change_majorcats)

expenditure_change_majorcats |> 
  kbl(caption = "Yearly Change in Expenditures", row.names = FALSE, align = "l") |> 
  kable_classic() |>
    row_spec(lastrow, bold = T, color = "black", background = "gray")

```

Export summary file with totals:

```{r exportexcelfile, warning = TRUE, message = TRUE}
# Saves main items in one excel file named `summary_file.xlsx`. Delete `eval=FALSE` to run on local computer.

#install.packages("openxlsx")
library(openxlsx)

#aggregated_totals_majorcats = rbind(rev_long, exp_long)

todaysdate = Sys.Date()
dataset_names <- list(#'Aggregate Revenues' = revenue_wide2, 
                      #'Aggregate Expenditures' = expenditure_wide2, 


                      
                      'Table 1' = revenue_change_majorcats, #Top categories with yearly change, 23 yr cagr
                      'Table 2' = expenditure_change_majorcats,
                      
                      'Appendix 1' = revenue_change2,
                      'Appendix 2' = expenditure_change2,
                      
                      'CAGR Rev-MajorCats' = CAGR_revenue_majorcats_tot, # Categories Match Table 1 in paper
                      'CAGR Exp-MajorCats' = CAGR_expenditures_majorcats_tot,  
                                            
                      'Fiscal Gap' = year_totals,    # Total Revenue, Expenditure, and Fiscal gap per year
                      
                      'aggregated_totals_long' = aggregated_totals_long#, # all data in long format. Good for creating pivot tables in Excel
                      
                   #   'aggregated_fewercategories' = aggregated_totals_majorcats # combines categories that have small amounts into "All Other" category
                      )

# no pension categories included as separate category (except for specific pension repayments that happened in some years.) 
# pension costs are included with K-12 education, university education, etc
write.xlsx(dataset_names, file = paste0("../Fiscal-Future-Topics/data/FY", current_year, " Files/summary_file_FY", current_year,"_", todaysdate, "_withpensioncat", ".xlsx"))
```


```{r  export-datahub-downloads}
library(openxlsx)

wb <- createWorkbook()


# ---------- Expenditures sheet ----------
addWorksheet(wb, "Expenditures")

writeDataTable(
  wb,
  "Expenditures",
  datahub_exp,
  tableStyle = "TableStyleMedium2"
)

col_widths <- pmax(
  nchar(names(datahub_exp)),
  sapply(datahub_exp, function(x) max(nchar(as.character(x)), na.rm = TRUE))
)

setColWidths(
  wb,
  "Expenditures",
  cols = 1:ncol(datahub_exp),
  widths = col_widths + 3
)


# ---------- Revenues sheet ----------
addWorksheet(wb, "Revenues")

writeDataTable(
  wb,
  "Revenues",
  datahub_rev,
  tableStyle = "TableStyleMedium2"
)

col_widths <- pmax(
  nchar(names(datahub_rev)),
  sapply(datahub_exp, function(x) max(nchar(as.character(x)), na.rm = TRUE))
)

setColWidths(
  wb,
  "Revenues",
  cols = 1:ncol(datahub_rev),
  widths = col_widths + 3
)

# ---------- README sheet ----------
addWorksheet(wb, "README")


readme_text <- data.frame(
  README = c(
    "Fiscal Futures DataHub Download",
    "",
    paste0("Last updated: ", todaysdate),
    "",
    "This file contains data downloaded from the Institute of Government and Public Affairs DataHub:",
    "https://igpa.uillinois.edu/igpa-data-hub",
    "",
    "Dollar values are in 1000s of nominal dollars",
    "",
    "Full documentation is available on GitHub:",
    "https://github.com/igpa-uillinois/Fiscal-Futures",
    "",
    "Dataset design and construction:",
    "Professor David Merriman (University of Illinois Chicago)",
    "Email: dmerrim@uic.edu"
  )
)

writeData(wb, "README", readme_text, colNames = FALSE)

setColWidths(wb, "README", 1, 110)

wrap <- createStyle(wrapText = TRUE)
addStyle(wb, "README", wrap, rows = 1:nrow(readme_text), cols = 1, gridExpand = TRUE)


# ---------- Save ----------
saveWorkbook(
  wb,
  paste0(
    "../Fiscal-Future-Topics/data/FY",
    current_year,
    " Files/IGPA_Datahub_Download",
    "_",
    todaysdate,
    "_withpensioncat",
    ".xlsx"
  ),
  overwrite = TRUE
)
```

```{r}
#| eval: false
#| include: false
#| 

## left for reference in case anyone wants it without formatting.
## Unformatted version of excel file
readme_text <- data.frame(
  README = paste(
    "This file contains data downloaded from the Institute of Government and Public Affairs at the University of Illinois's DataHub",
    "https://igpa.uillinois.edu/igpa-data-hub",
    "",
    "Dollar values are in 1000s of nominal dollars",
    "",
    "The data is fully documented on the Github for the project",
    "https://github.com/igpa-uillinois/Fiscal-Futures",
    "",
    "Professor David Merriman of the University of Illinois Chicago is responsible for the design and construction of the data set.",
    "You may contact him at dmerrim@uic.edu",
    sep = "\n"
  )
)


dataset_names <- list(
  'Expenditures' = datahub_exp,
  'Revenues'     = datahub_rev,
  'README'       = readme_text
                      )

write.xlsx(dataset_names, file = paste0("../Fiscal-Future-Topics/data/FY", current_year, " Files/datahub_download", current_year,"_", todaysdate, "_withpensioncat", ".xlsx"))

```

